---
title: "Tidymodels Workflows with yaap"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Tidymodels Workflows with yaap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
set.seed(42L)
par(bty = "n")
```

## Overview

yaap fits archetypal analysis (AA) models. In a tidymodels workflow,
it is best understood as an unsupervised decomposition or feature-extraction
tool, closer in role to `stats::prcomp()` than to a supervised `parsnip`
model specification. yaap learns archetype coordinates and sample
composition weights, then exposes those results through interfaces that work
naturally with the tidymodels ecosystem:

* `step_archetypes()` is a `recipes` step that learns archetypes during
  `prep()` and returns archetype composition weights during `bake()`.
* `tune_archetypes()` fits a grid of AA models over candidate values of `K`, the
  number of archetypes.

This vignette uses only lightweight pieces of the ecosystem. Because AA is an
unsupervised representation method, the examples focus on preprocessing,
feature extraction, and model inspection rather than on `parsnip` fitting.

```{r libraries, message=FALSE, warning=FALSE}
library(yaap)
library(recipes)
library(tune)
```

The examples use the small two-dimensional toy data included with yaap.

```{r plot-toy, fig.cap = "The toy data contain three visually apparent extremes."}
toy <- read.csv(system.file("extdata", "toy.csv", package = "yaap"))
plot(
  toy,
  asp = 1,
  pch = 19,
  cex = 0.6,
  col = "grey50",
  main = "Toy data",
  frame.plot = FALSE
)
```

## Choosing K with scree, elbow, and AIC

The main tuning parameter in archetypal analysis is `K`: the number of
archetypes. A traditional exploratory approach, also familiar from the classic
`archetypes` package, is to fit a sequence of models and inspect a screeplot or
elbow plot. The elbow is the smallest `K` after which reconstruction quality
improves only modestly.

yaap also provides an adapted AIC-like criterion for Euclidean Gaussian AA
models. This is a yaap-specific model-selection helper; it is not part of
the classic `archetypes` package workflow. It is useful for comparing candidate
`K` values on the same data, but it is still a model-selection aid rather
than a substitute for scientific judgment.

Here we fit `K = 1, ..., 6`. The one-archetype baseline is useful for RSS and
R2, but the adapted AIC can be undefined for `K = 1`, so we score the grid by
RSS and compute AIC separately where it is defined.

```{r k-selection-fit}
rss <- function(fit) {
  utils::tail(fit$loss$loss, 1L)
}

aic_or_na <- function(fit) {
  if (nrow(fit$coordinates) == 1L) NA_real_ else AIC(fit)
}

set.seed(42L)
ens_k <- tune_archetypes(
  toy,
  K = 1:6,
  nrep = 1L,
  init = "furthest_sum",
  max_iter = 30L,
  tol_r2 = 0,
  max_kappa = Inf,
  eval = rss,
  eval_name = "rss"
)

k_tbl <- summary(ens_k)
k_tbl$AIC <- vapply(ens_k$fits, aic_or_na, numeric(1L))
knitr::kable(
    k_tbl[, c("K", "rss", "r2", "AIC")],
    digits = c(0, 1, 3, 1),
)
```

```{r k-selection-plot, fig.width = 10.5, fig.height = 3.7, fig.cap = "RSS, R-squared, and adapted AIC across candidate values of K."}
op <- par(mfrow = c(1, 3), mar = c(4, 4, 2.5, 1), bty = "n")

plot(
  k_tbl$K, k_tbl$rss,
  type = "b",
  pch = 19,
  frame.plot = FALSE,
  xlab = "Number of archetypes (K)",
  ylab = "Reconstruction loss",
  main = "RSS"
)
abline(v = 3, lty = 2, col = "grey50")

plot(
  k_tbl$K, 1 - k_tbl$r2,
  type = "b",
  pch = 19,
  frame.plot = FALSE,
  xlab = "Number of archetypes (K)",
  ylab = expression(R^2),
  main = "Unexplained Variance (1 - R2)"
)
abline(v = 3, lty = 2, col = "grey50")

plot(
  k_tbl$K, k_tbl$AIC,
  type = "b",
  pch = 19,
  frame.plot = FALSE,
  xlab = "Number of archetypes (K)",
  ylab = "Adapted AIC",
  main = "Information Criterion"
)
abline(v = 3, lty = 2, col = "grey50")

par(op)
```

The RSS and R2 curves show that `K = 3` is already a strong low-dimensional
summary, but they do not make the choice completely unambiguous: larger models,
especially `K = 5`, can look roughly comparable by reconstruction quality. In
this example, yaap's adapted AIC is the most decisive diagnostic, giving its clear
minimum at `K = 3`.

```{r best-model}
selected_ix <- which.min(k_tbl$AIC)
selected_K <- k_tbl$K[selected_ix]
fit <- ens_k$fits[[selected_ix]]

k_tbl[selected_ix, c("K", "rss", "r2", "AIC")]
selected_K
```

In real analyses these plots should be treated as diagnostics, not as automatic
rules. AA optimization is non-convex, so single-restart curves can be noisy and
may not decrease monotonically for every larger `K`. Use more restarts and
domain knowledge when the elbow or criterion is ambiguous.


## Using archetypes in recipes

`step_archetypes()` turns AA into a preprocessing step. During `prep()`, the
step fits archetypes on the selected numeric predictors. During `bake()`, it
projects data onto the learned archetype simplex and returns one composition
column per archetype.

The recipes argument is called `num_comp`, following tidymodels naming
conventions, but it is the same quantity as `K`.

```{r recipe-fit}
rec <- recipe(~ ., data = toy) |>
  step_archetypes(
    all_numeric(),
    num_comp = selected_K,
    seed = 42L,
    options = list(
      init = "furthest_sum",
      max_iter = 30L,
      tol_r2 = 0,
      max_kappa = Inf
    )
  )

rec_prep <- prep(rec, training = toy)
rec_prep
```

```{r recipe-bake}
baked <- bake(rec_prep, new_data = toy)
head(baked)

range(rowSums(as.matrix(baked[, anames(fit)])))
```

By default, the original numeric predictors are removed and replaced by
composition columns named `A1`, `A2`, and so on. Set
`keep_original_cols = TRUE` when both the original predictors and archetype
weights should be retained.

```{r recipe-keep-original}
rec_keep <- recipe(~ ., data = toy) |>
  step_archetypes(
    all_numeric(),
    num_comp = selected_K,
    keep_original_cols = TRUE,
    seed = 42L,
    options = list(max_iter = 30L, tol_r2 = 0, max_kappa = Inf)
  ) |>
  prep(training = toy)

head(bake(rec_keep, new_data = toy))
```

## Comparing `step_archetypes()` with `step_pca()`

The closest tidymodels analogy for yaap is `recipes::step_pca()`: both steps
learn an unsupervised representation of the numeric predictors during `prep()`.
The outputs have different meanings. PCA returns orthogonal score columns such
as `PC1` and `PC2`; AA returns composition weights such as `A1`, `A2`, and `A3`
that are non-negative and sum to one for each sample.

The two recipes below are trained in parallel on the same toy data.

```{r compare-pca-aa}
rec_aa <- recipe(~ ., data = toy) |>
  step_archetypes(
    all_numeric(),
    num_comp = 3L,
    seed = 42L,
    options = list(
      init = "furthest_sum",
      max_iter = 30L,
      tol_r2 = 0,
      max_kappa = Inf
    )
  ) |>
  prep(training = toy)

rec_pca <- recipe(~ ., data = toy) |>
  step_pca(all_numeric(), num_comp = 2L) |>
  prep(training = toy)

aa_scores <- bake(rec_aa, new_data = toy)
pca_scores <- bake(rec_pca, new_data = toy)

head(aa_scores)
head(pca_scores)
```

The raw AA columns live on a simplex, so they are bounded and sum to one. If
you want unbounded Euclidean coordinates, use a log-ratio transformation. The
centered log-ratio (CLR) transformation treats the archetypes symmetrically by
subtracting each row's mean log weight. The additive log-ratio (ALR)
transformation instead selects a base component; here we use `A1` as the base
and plot `log(A2 / A1)` against `log(A3 / A1)`.

```{r compare-pca-aa-logratio}
aa_cols <- c("A1", "A2", "A3")
aa_mat <- as.matrix(aa_scores[, aa_cols])
aa_mat <- pmax(aa_mat, 1e-6)
aa_mat <- aa_mat / rowSums(aa_mat)

log_aa <- log(aa_mat)
aa_clr <- log_aa - rowMeans(log_aa)
colnames(aa_clr) <- paste0("clr_", aa_cols)

aa_alr <- data.frame(
  log_A2_A1 = log(aa_mat[, "A2"] / aa_mat[, "A1"]),
  log_A3_A1 = log(aa_mat[, "A3"] / aa_mat[, "A1"])
)

head(aa_clr)
head(aa_alr)
```

```{r compare-pca-aa-plot, fig.width = 7, fig.height = 3.5, fig.cap = "PCA scores are variance-coordinate projections; ALR-transformed AA scores are log-ratio coordinates of archetype compositions."}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1), bty = "n")
alr_atan <- with(aa_alr, atan2(log_A3_A1 - mean(log_A3_A1), log_A2_A1 - mean(log_A2_A1)))
col_pt <- colorRampPalette(c("blue", "gray", "red"))(length(alr_atan))[order(order(alr_atan))]

plot(
  pca_scores$PC1, pca_scores$PC2,
  pch = 19,
  cex = 0.6,
  col = col_pt,
  xlab = "PC1",
  ylab = "PC2",
  main = "step_pca()",
  frame.plot = FALSE
)
abline(h = 0, v = 0, col = "grey85")

plot(
  aa_alr$log_A2_A1, aa_alr$log_A3_A1,
  pch = 19,
  cex = 0.6,
  col = col_pt,
  xlab = "log(A2 / A1)",
  ylab = "log(A3 / A1)",
  main = "step_archetypes() + ALR",
  frame.plot = FALSE
)
abline(h = 0, v = 0, col = "grey85")

par(op)
```

The PCA plot is a rotated coordinate system for the original variables. The AA
plot starts from mixture weights over interpretable extremes, then uses a
log-ratio transform to remove the simplex constraint. CLR is the symmetric
choice when no archetype should be privileged; ALR is the base-selected version
shown above and gives `K - 1` unconstrained coordinates.

## Declaring tunable parameters

`step_archetypes()` declares `num_comp` and `delta` as tunable recipe
parameters. This lets the step participate in a larger tidymodels tuning
workflow when the rest of the modeling pipeline supplies resampling, metrics,
and an outcome.

```{r tunable}
rec_tune <- recipe(~ ., data = toy) |>
  step_archetypes(
    all_numeric(),
    num_comp = tune(),
    delta = tune()
  )

tunable(rec_tune$steps[[1L]])
```

For unsupervised AA exploration, `tune_archetypes()` is often the most direct
tool. For supervised modeling pipelines, use `step_archetypes()` inside the
broader tidymodels workflow and tune `num_comp` alongside the downstream model
parameters.
