---
title: "Choosing an Initialization Method for Archetypal Analysis"
bibliography: ../references.bib
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Choosing an Initialization Method for Archetypal Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5.5,
    fig.align = "center",
    warning = FALSE
)
set.seed(42)
par(bty = "n")
library(yaap)
```

> This vignette assumes familiarity with the basic AA workflow covered in the
  **Introduction to Archetypal Analysis with yaap** vignette.

## Quick Start

For a first Euclidean AA fit, start with `"furthest_sum"` or `"kmeans_pp"`.
`"furthest_sum"` is a strong boundary-seeking default for convex data, while
`"kmeans_pp"` is a softer alternative when you want some randomness without
falling back to uniform sampling. If the geometry is unknown and you can afford
several starts, use `"random"` with a larger `nrep`. If you only have a few
starts and runtime is less important, use `"aa_pp"` because each added point is
chosen against the current AA reconstruction error.

```{r quick-recommendation, eval = FALSE}
library(yaap)

toy <- read.csv(system.file("extdata", "toy.csv", package = "yaap"))

fit <- run_aa(toy, K = 3, nrep = 5,  init = "furthest_sum")
fit <- run_aa(toy, K = 3, nrep = 5,  init = "kmeans_pp")
fit <- run_aa(toy, K = 3, nrep = 20, init = "random")
fit <- run_aa(toy, K = 3, nrep = 3,  init = "aa_pp")
```

## Background

Archetypal analysis (AA) minimizes a reconstruction loss, written in the
standard Euclidean formulation as

$$\mathcal{L}(X, SA) = \|X - S A\|_F^2.$$

As with all the matrix decomposition problems, the objective is non-convex in
both $S$ and $A$, and gradient descent can terminate at a **local minimum**. As
a result, the quality of the final solution depends substantially on the starting
point.

yaap implements several options for initializing the archetype coordinates,
which have been suggested in the literature, through the `aa_init()` function
and the `init` argument.

## Methods at a Glance

The table below summarizes the seven methods by their strategy, whether they
involve random sampling, their relative speed, and any mandatory extra arguments.

| Method                 | Strategy                                                                           | Stochastic?      | Speed  | Mandatory extras |
| ---------------------- | ---------------------------------------------------------------------------------- | ---------------- | ------ | :--------------: |
| `"random"`             | Sample $K$ rows uniformly at random                                                | Yes              | Fast   | —                |
| `"dirichlet"`          | Construct each archetype as a random convex combination of rows                    | Yes              | Fast   | —                |
| `"kmeans_pp"`          | Sample proportional to squared distance to the nearest current archetype           | Yes              | Medium | —                |
| `"aa_pp"`              | Sample using residual reconstruction error from the current hull                   | Yes              | Slow   | —                |
| `"furthest_first"`     | Select the point furthest from all current archetypes                              | Seed only        | Medium | —                |
| `"furthest_sum"`       | Select the point that maximizes the *sum* of distances to all current archetypes   | Seed only        | Medium | —                |
| `"hull_outmost"`       | Select frequently outlying convex-hull candidates                                  | Partitioned only | Medium | `hull_method`    |

Large-data approximations are controlled orthogonally with `batch_size`,
`batch_type`, and `batch_replace`. By default, `batch_type = "distal"` samples
candidate rows with probability proportional to squared distance from the data
center, preserving the coreset idea of focusing computation on likely boundary
points [@Mair2019]. Use `batch_type = "uniform"` when you want an unbiased
mini-batch approximation, for example with `method = "aa_pp"`.


## The `aa_init()` Function

`aa_init()` runs any initialization method as a standalone step, returning a
named list with two elements:

* **`A`** — $K \times M$ matrix of initial archetype coordinates.
* **`B`** — $K \times N$ row-stochastic matrix expressing each archetype as a
  convex combination of data points ($A = B X$).

```{r aa-init-basic}
toy <- read.csv(system.file("extdata", "toy.csv", package = "yaap"))
X <- as.matrix(toy) # 250 × 2
K <- 3

init <- aa_init(X, K = K, method = "furthest_sum")
str(init)
```

The `A` matrix provides the initial archetype coordinates that will seed the
optimization loop; `B` is the convex-hull certificate for those coordinates. If
`run_aa()` is initialized with `aa_init()` and a method name or a custom function,
the `B` returned is carried into the algorithm initialization. If you instead
pass a pre-computed coordinate matrix such as `init$A`, `run_aa()` checks that
the coordinates still admit a feasible `B`. Rows outside the allowed data hull
are projected back into the hull with a warning.

```{r initialization-helpers, include = FALSE}
draw_simplex <- function(A, border, fill = NULL, lty = 1, lwd = 2,
                         pch = 17, cex = 1.1) {
    if (nrow(A) >= 3) {
        hull <- chull(A)
        polygon(A[hull, , drop = FALSE],
            border = border, lwd = lwd, lty = lty,
            col = fill
        )
    } else if (nrow(A) == 2) {
        lines(A, col = border, lwd = lwd, lty = lty)
    }
    points(A, pch = pch, cex = cex, col = border)
}

selection_deviation_score <- function(counts, K, n, pseudocount = 0.5) {
    total_counts <- sum(counts)
    N_runs <- total_counts / K
    expected <- total_counts / n
    p_null <- K / n

    log_fc <- log2((counts + pseudocount) / (expected + pseudocount))
    p_lower <- pbinom(counts, size = N_runs, prob = p_null)
    p_upper <- pbinom(counts - 1, size = N_runs, prob = p_null,
        lower.tail = FALSE
    )
    p_value <- pmin(1, 2 * pmin(p_lower, p_upper))
    signed_log_p <- sign(counts - expected) *
        -log10(pmax(p_value, .Machine$double.xmin))

    list(log_fc = log_fc, p_value = p_value, signed_log_p = signed_log_p)
}

plot_selected_density <- function(X, counts, main, K, color_max = NULL,
                                  pseudocount = 0.5) {
    score <- selection_deviation_score(counts, K, nrow(X), pseudocount)
    signed_log_p <- score$signed_log_p

    pal <- colorRamp(c("blue", "white", "red"))
    cmax <- color_max
    if (is.null(cmax)) cmax <- max(abs(signed_log_p))
    probs <- if (cmax > 0) {
        pmax(0, pmin(1, (signed_log_p + cmax) / (2 * cmax)))
    } else {
        rep(0.5, length(signed_log_p))
    }
    point_cols <- pal(probs) / 255
    point_cols <- rgb(point_cols[, 1], point_cols[, 2], point_cols[, 3])
    ix <- order(abs(signed_log_p))
    plot(
        X[ix, ],
        pch = 21, cex = 1, col = "gray50", bg = point_cols[ix],
        main = main, xlab = "", ylab = "", axes = FALSE, asp = 1
    )
}

plot_data_with_points <- function(X, A, main, col = "grey75", view = c(1, 2),
                                  pch = 4, point_col = "black") {
    plot(
        X[, view, drop = FALSE],
        pch = 16, cex = 0.45, col = col,
        main = main, xlab = "", ylab = "", axes = FALSE, asp = 1
    )
    points(A[, view, drop = FALSE],
        pch = pch, cex = 1.25, lwd = 1.6,
        col = point_col
    )
}

make_concentric_circles <- function(n = 180, noise = 0.03) {
    n_inner <- n %/% 2
    n_outer <- n - n_inner
    theta   <- c(runif(n_inner, 0, 2 * pi),
                 runif(n_outer, 0, 2 * pi))
    radius  <- c(rep(0.4, n_inner),
                 rep(1.0, n_outer))
    X <- cbind(radius * cos(theta),
               radius * sin(theta))
    X <- X + rnorm(length(X), sd = noise)
    colnames(X) <- c("x", "y")
    list(
        X = X,
        view = c(1, 2),
        col = c(rep("#0072B2", n_inner), rep("#D55E00", n_outer)),
        main = "Concentric circles",
        kernel = "laplace",
        kernel_args = list(sigma = 0.25)
    )
}

make_swiss_roll <- function(n = 180, noise = 0.025) {
    t <- runif(n, 1.5 * pi, 4.5 * pi)
    h <- runif(n, -1, 1)
    X <- cbind(t * cos(t) / 7, h, t * sin(t) / 7)
    X <- X + rnorm(length(X), sd = noise)
    colnames(X) <- c("x", "height", "z")
    roll_cols <- colorRampPalette(c("#0072B2", "#009E73", "#D55E00"))(100)
    col_id <- pmax(1, pmin(100, ceiling(99 * (t - min(t)) / diff(range(t))) + 1))
    list(
        X = X,
        view = c(1, 3),
        col = roll_cols[col_id],
        main = "Swiss roll",
        kernel = "rbf",
        kernel_args = list(sigma = 0.45)
    )
}
```

### Supplying Your Own Initialization

The built-in method names cover the common cases, but `run_aa()` also accepts
either a precomputed matrix of archetype coordinates or a custom initialization
function.

If the starting coordinates are already available, pass them directly as a
matrix in the same units as the `X`. `run_aa()` will preprocess it the same way
it preprocesses `X`, handles the required checks, and constructs `B`.

```{r precomputed-init-matrix}
X <- as.matrix(toy)
K <- 3
tol <- 0.01
eps <- 0

A0 <- X[1:K, , drop = FALSE]

fit_matrix <- run_aa(X, K = K, init = A0, scale = FALSE, tol = tol, eps = eps)
```

A custom function is useful when initialization itself needs code. The function
must take `X` and `K`, and return the same two objects as `aa_init()`: the
archetype coordinates `A` and the data weights `B`. Additional arguments can
be passed through `init_args`. Below is a deliberately simple initializer that
chooses the first `K` rows of the data.

```{r custom-init-function}
X <- as.matrix(toy)
K <- 3
tol <- 0.01

first_k_init <- function(X, K, ...) {
    A <- X[1:K, , drop = FALSE]
    B <- diag(1, nrow = K, ncol = nrow(X))
    list(A = A, B = B)
}

fit_custom <- run_aa(X, K = K, init = first_k_init, scale = FALSE, tol = tol)
```

If your custom initializer naturally produces only archetype coordinates,
`fit_simplex()` exposes the same simplex mapping used internally to construct a
feasible `B`. Reconstructing `A_init` as `B %*% X` ensures the coordinates passed
to AA are on the data hull:

```{r custom-init-fit-simplex, eval = FALSE}
X <- as.matrix(toy)
K <- 3

A_user <- some_initializer(X, K = K)
B <- fit_simplex(A_user, X)
A_init <- B %*% X
init <- list(A = A_init, B = B)
```

These two forms are useful when the starting point comes from a previous
analysis, a domain-specific rule, or an external clustering workflow. For most
workflows, passing a method name remains the shortest form.

```{r run-aa-init-short, eval = FALSE}
X <- as.matrix(toy)
K <- 3
tol <- 0.01

fit <- run_aa(X, K = K, init = "furthest_sum", tol = tol)
fit <- run_aa(X, K = K, init = "aa_pp", tol = tol)
fit <- run_aa(X,
    K = K, init = "aa_pp",
    init_args = list(batch_size = 200, batch_type = "uniform"),
    tol = tol
)
```

## Comparing Initializer Biases

```{r init-selection-density}
X <- as.matrix(toy)
K <- 3

method_labels <- c(
    random         = "Random",
    kmeans_pp      = "k-means++",
    aa_pp          = "AA++",
    furthest_first = "Furthest First",
    furthest_sum   = "Furthest Sum",
    hull_outmost   = "Hull-outmost"
)
N_expected <- 10  # Expected selections per point across all runs
N_runs <- nrow(X) * N_expected
# "significance" cutoff for coloring points red/blue
color_cutoff <- -log10(1 / nrow(X))

selected_counts <- list()
for (method in names(method_labels)) {
    counts <- rep(0, nrow(X))
    for (run in 1:N_runs) {
        if (method == "hull_outmost") {
            init_obj <- aa_init(
                X, K = K,
                method = method,
                hull_method = "partitioned"
            )
        } else {
            init_obj <- aa_init(X, K = K, method = method)
        }
        selected <- which(init_obj$B > 0, arr.ind = TRUE)[, 2]
        counts[selected] <- counts[selected] + 1
    }
    selected_counts[[method]] <- counts
}
```

The first practical difference between initializers is not whether they are
"good" or "bad"; it is where they tend to look. The toy data below are simple:
250 two-dimensional points spread near the vertices of a triangle. The figure
below repeats each row-selecting initializer many times and colors each point by
how often it was selected compared with uniform random sampling. Red points are
selected more often than random, blue points are selected less often, and white
points are close to the random baseline.

```{r init-selection-density-plot, fig.width = 8, fig.height = 7, out.width="100%", results='hold', echo=FALSE}

op <- par(mfrow = c(2, 3), mar = c(2, 2, 3, 0.5), bty = "n")
for (method in names(method_labels)) {
    plot_selected_density(
        X,
        selected_counts[[method]],
        main = method_labels[method],
        K = K,
        color_max = color_cutoff
    )
}
par(op)
```

As expected, `random` selects archetypes uniformly from the observed rows, so
the points do not diverge much from their null probability of selection. This
is probably not the best strategy for a convex geometry like the toy data, but
it is unbiased and thus robust to unknown data structure. With enough restarts
this strategy will eventually explore every region, but any single run may miss
the corners.

On the other extreme, methods like `furthest_first`, `furthest_sum`, and
`hull_outmost` are exploring more aggressively the
boundary of the data geometry. We see that most of them converge to the same set
of candidate points and never pick any point in the middle of the cloud.
`furthest_first` is the only one that shows some variation in the selected
points, but it still has a strong corner preference. For a simple convex shape
that bias is often useful: the corners are exactly where good archetypes should
start.

Finally, in the middle between the two extremes, are clustering-inspired methods
like: `kmeans_pp` and `aa_pp`. These methods adapt ideas developed
to initialize clustering algorithms to the AA setting. They are more probabilistic
in nature and thus represent softer versions of their boundary searching counterparts.
They also rarely select any point in the center of the cloud but they are exploring
more diffusely the boundary regions and are not actively looking for the extrema.

More details about the methodology employed by each method can be found in the
[Method Reference](#method-reference) section at the end of this vignette.
The rest of the vignette narrows to three representatives: `random` as an
exploratory baseline, `kmeans_pp` as the softer outward-biased initializer, and
`furthest_sum` for the strong greedy boundary bias.

## Increasing the Number of Archetypes

With $K = 3$, the toy data roughly match the three-corner shape. When $K$
increases, the question changes from "did we find the corners?" to "how do the
extra archetypes cover the cloud?" The following chunk is the user-facing call:
only the `method` and `K` values change.

```{r increasing-k}
X <- as.matrix(toy)
k_grid <- c(3, 7, 15)
k_methods <- c("random", "kmeans_pp", "furthest_sum")

k_inits <- list()
for (method in k_methods) {
    method_inits <- list()
    for (K in k_grid) {
        k_lab <- paste("K =", K)
        method_inits[[k_lab]] <- aa_init(X, K = K, method = method)
    }
    k_inits[[method]] <- method_inits
}
```

```{r increasing-k-plot, fig.width = 8.5, fig.height = 6, out.width="100%", results='hold', echo=FALSE}
op <- par(
    mfrow = c(length(k_methods), length(k_grid)),
    mar = c(0.5, 0.5, 2.2, 0.2), oma = c(0, 0, 0, 0), bty = "n"
)
for (method in k_methods) {
    for (k_label in names(k_inits[[method]])) {
        init_obj <- k_inits[[method]][[k_label]]
        plot(
            X,
            pch = 16, cex = 0.42, col = "grey75",
            main = sprintf("%s\n%s", method_labels[method], k_label),
            xlab = "", ylab = "", axes = FALSE, asp = 1
        )
        draw_simplex(
            init_obj$A,
            border = "firebrick3",
            fill = adjustcolor("firebrick3", 0.05),
            cex = 0.95
        )
    }
}
par(op)
```

As $K$ grows, the non-aggressive methods start to waste extra archetypes in
the center of the convex hull. This is visible for `random`, and eventually also
for `kmeans_pp`: once the main corners have been sampled, additional points are
often placed in the interior regions where there are definitely no archetypes
for this triangle. Those starts can still optimize successfully, but the
initialization budget is no longer being used to approximate the boundary.

`furthest_sum` behaves differently. It keeps adding points on the outside of the
cloud and, as $K$ increases, the selected polygon becomes a finer approximation
to the true convex hull boundary. In fact, `furthest_sum` generated points
are "guaranteed to lie in the minimal convex set of the unselected data points"
[Theorem 2 in [@Morup2012]]

## When Initialization Matters

The toy triangle is intentionally simple. If we start from one initialization
and then let `run_aa()` optimize, most methods reach essentially the same final
fit. The red dashed triangle below is the starting point; the blue solid
triangle is the fitted solution.

```{r init-before-after-fit}
X <- as.matrix(toy)
K <- 3
focused_methods <- c("random", "kmeans_pp", "furthest_sum")
fits <- list()
for (method in focused_methods) {
    fits[[method]] <- run_aa(
        x        = X,
        K        = K,
        init     = method,
        scale    = FALSE,
        max_iter = 60,
        tol      = 0.01
    )
}
```

```{r init-before-after-plot, echo=FALSE, fig.width = 8.5, fig.height = 6, out.width="100%", results='hold'}
loss_values <- c()
for (method in focused_methods) {
    loss_values <- c(loss_values, fits[[method]]$loss$loss)
}
loss_ylim <- range(loss_values)
op <- par(mfrow = c(2, 3), mar = c(2.2, 2.5, 3, 0.5), bty = "n")
for (method in focused_methods) {
    fit <- fits[[method]]
    plot(fit,
        what = "coordinates",
        show_anames = FALSE,
        main = sprintf("%s\n%d steps", method_labels[method], nrow(fit$loss) - 1),
        args.data.scatter = list(pch = 16, cex = 0.45, col = "grey75"),
        pch = 19,
        col = "steelblue4",
        axes = FALSE,
        asp = 1
    )
    draw_simplex(
        fit$init,
        border = adjustcolor("firebrick3", 0.75),
        fill = adjustcolor("firebrick3", 0.05),
        lty = 2
    )
}
for (method in focused_methods) {
    plot(fits[[method]],
        what = "loss",
        lwd = 1.8,
        col = "steelblue4",
        ylim = loss_ylim,
        main = "Loss",
        ylab = "",
        las = 1
    )
}
par(op)
```

This is the reassuring case: the data geometry is simple enough that a mediocre
start can still be corrected by the optimizer. Initialization matters more when
the geometry is less well described by straight-line Euclidean distances.

## Nonlinear Geometries

The previous examples still live in a simple Euclidean geometry: distances in
the plotted coordinates describe the structure we care about. Some datasets are
not like that. In concentric circles, the inner circle is important even though
the outer circle contains the most distant points. In a Swiss roll, points that
look close in a two-dimensional projection may be far apart along the rolled
surface.

```{r nonlinear-init}
K <- 8
nonlinear_data <- list(
    circles    = make_concentric_circles(),
    swiss_roll = make_swiss_roll()
)
nonlinear_methods <- c("random", "kmeans_pp", "furthest_sum")

nonlinear_inits <- list()
for (data_shape in names(nonlinear_data)) {
    data_obj <- nonlinear_data[[data_shape]]
    out <- list()
    for (method in nonlinear_methods) {
        out[[method]] <- aa_init(data_obj$X, K = K, method = method)
    }
    nonlinear_inits[[data_shape]] <- out
}
```

```{r nonlinear-init-plot, fig.width = 8.5, fig.height = 6, out.width="100%", results='hold', echo=FALSE}
op <- par(
    mfrow = c(length(nonlinear_data), length(nonlinear_methods)),
    mar = c(2, 2, 3, 0.5), bty = "n"
)
for (data_shape in names(nonlinear_data)) {
    data_obj <- nonlinear_data[[data_shape]]
    for (method in nonlinear_methods) {
        init_obj <- nonlinear_inits[[data_shape]][[method]]
        plot_data_with_points(
            data_obj$X,
            init_obj$A,
            main = sprintf("%s\n%s", data_obj$main, method_labels[method]),
            col = adjustcolor(data_obj$col, 0.62),
            view = data_obj$view
        )
    }
}
par(op)
```

First inspect the Euclidean initializations. The biased methods are doing what
they were designed to do: they seek far-away points. On these shapes, that can
over-emphasize the outside and leave the center or inner structure
under-represented. `random` does not fix the geometry, but it is less committed
to the outside by construction.

For these cases, the better answer is often to change the space in which AA is
fit. A kernel method lets nearby points influence each other according to a
smooth similarity rule rather than only through straight-line distance in the
input coordinates.

```{r kernel-user-code, eval=FALSE}
fit_kernel <- archetypes_kernel_pgd(
    x      = X,
    K      = K,
    kernel = "laplace",
    init   = "furthest_sum",
    tol    = 0.01
)
```

```{r nonlinear-kernel-setup, include = FALSE, warning=FALSE}
kernel_methods <- c("furthest_sum", "kmeans_pp")
kernel_inits <- list()
for (data_shape in names(nonlinear_data)) {
    data_obj <- nonlinear_data[[data_shape]]
    out <- list()
    for (method in kernel_methods) {
        kernel_fit <- archetypes_kernel_pgd(
            data_obj$X,
            K = K,
            init = method,
            kernel = data_obj$kernel,
            kernel_args = data_obj$kernel_args,
            max_iter = 1,
            tol = 0.01,
            tol_r2 = 1
        )
        out[[method]] <- list(
            euclidean = nonlinear_inits[[data_shape]][[method]]$A,
            kernel = kernel_fit$init %*% data_obj$X
        )
    }
    kernel_inits[[data_shape]] <- out
}
```

```{r nonlinear-kernel-plot, fig.width = 8.5, fig.height = 6, out.width="100%", results='hold', echo=FALSE}
op <- par(
    mfrow = c(length(nonlinear_data), length(kernel_methods)),
    mar = c(2, 2, 3, 0.5), bty = "n"
)
for (data_shape in names(nonlinear_data)) {
    data_obj <- nonlinear_data[[data_shape]]
    for (method in kernel_methods) {
        init_obj <- kernel_inits[[data_shape]][[method]]
        plot(
            data_obj$X[, data_obj$view, drop = FALSE],
            pch = 16, cex = 0.45, col = adjustcolor(data_obj$col, 0.55),
            main = sprintf("%s\n%s", data_obj$main, method_labels[method]),
            xlab = "", ylab = "", axes = FALSE, asp = 1
        )
        points(init_obj$euclidean[, data_obj$view, drop = FALSE],
            pch = 4, cex = 1.2, lwd = 1.5, col = "black"
        )
        points(init_obj$kernel[, data_obj$view, drop = FALSE],
            pch = 1, cex = 1.25, lwd = 1.6, col = "#D55E00"
        )
    }
}
par(fig = c(0, 1, 0, 1), mar = c(0, 0, 0, 0), new = TRUE)
plot.new()
legend(
    "center",
    legend = c("Euclidean init", "Kernel init"),
    pch = c(4, 1), col = c("black", "#D55E00"),
    pt.cex = c(2.1, 2.2), bty = "n", cex = 1.2
)
par(op)
```

The same initialization names still work. The important difference is that the
kernel version measures spread in the similarity space used by the model. This
can rescue boundary-seeking methods on shapes where the raw coordinates make
the outside look more important than the rest of the structure.

## Conclusions and Recommendations

There is no universal best initializer. The right choice depends on how much
you know about the geometry.

| Situation                             | Recommended `init`                                                          |
| ------------------------------------- | ----------------------------------------------------------------------------|
| Simple convex-ish data                | `"furthest_sum"`, `"kmeans_pp"`                                             |
| Unknown structure, enough restarts    | `"random"` with a larger `nrep`                                             |
| Unknown structure, few restarts       | `"aa_pp"`                                                                   |
| Very large dataset, speed critical    | `"furthest_sum"` with `batch_size` or `"hull_outmost"` partitioned          |
| Reproducing legacy results            | `"furthest_sum"` (PCHA), `"random"` [archetypes], `"dirichlet"` directional |

When the geometry is simple, a boundary or coverage bias can be helpful. When
the geometry is unknown, `random` with enough restarts is often the safest
baseline because it makes the fewest assumptions.

If the geometry is visibly nonlinear, changing the model is usually more
important than fine-tuning the Euclidean initializer. Use
`archetypes_kernel_pgd()` with a suitable kernel, then choose one of the
kernel-compatible initializers such as `"kmeans_pp"` or `"furthest_sum"`.

For memory efficiency and speed, any initializer can be restricted to a
candidate batch with `batch_size`. The default `batch_type = "distal"` samples
candidate rows in proportion to their squared distance from the data center,
which is the coreset idea of @Mair2019: if archetypes are expected near the
boundary, it is wasteful to spend most candidate memory on central points. Use
`batch_type = "uniform"` when the goal is an unbiased mini-batch approximation,
especially for `method = "aa_pp"`.

### Initialization Caveats {#initialization-caveats}

The default `scale = TRUE` in `run_aa()` applies column-wise standardization
(z-scoring) before fitting. While scaling often improves the conditioning of the
optimization, it interacts differently with each initialization strategy.

@Mair2023 show that FurthestSum is sensitive to preprocessing: standardization
can distort the Euclidean distances it uses to choose boundary points. AA++ is
less affected because it samples from residuals tied directly to the AA objective
[@Mair2023].

The cost of this objective-aware sampling is runtime. Among the built-in
initializers, `"aa_pp"` is usually the slowest because after the first two
points it solves a small AA projection problem at each step, essentially a
greedy AA analysis. In return, each added point is guaranteed not to increase
the reconstruction cost, and reduces it whenever it expands the current hull.

Initializer names are not supported identically by every model family. The
Euclidean solvers (`method = "pgd"`, `"nnls"`, and `"paa"`) call
`aa_init()` on the preprocessed data, so they accept the full catalogue of
initializer strings described above, as well as custom initializer functions and
coordinate matrices.

Kernel AA is the main exception. With `method = "kernel"`, initializer strings
are limited to `"random"`, `"dirichlet"`, `"furthest_first"`, `"kmeans_pp"`,
and `"furthest_sum"`. Row-selection methods can be expressed using selected row
indices and pairwise distances from the kernel Gram matrix; `"dirichlet"` is
also available because it constructs the coefficient matrix $B$ directly and
does not require input-space archetype coordinates. Candidate batching is
supported for these kernel initializers, with distal batches computed from
kernel-space center distances. The remaining strings are not accepted as kernel
initializers: `"aa_pp"` requires convex-hull residual projections and
`"hull_outmost"` requires explicit hull candidates. Those operations would need
separate kernel-space implementations. If you need one of those starts for a
kernel fit, pass an explicit `K x n` non-negative coefficient matrix instead;
row indices or row names are also accepted when you want to select observed rows
directly.

The directional solver also accepts the `aa_init()` method
strings, but it defaults to `"dirichlet"` for historical reasons.

## Method Reference {#method-reference}

### `random`

Selects $K$ rows of $X$ uniformly at random without replacement.

```{r uniform-ex, eval = FALSE}
X <- as.matrix(toy)
K <- 3
init <- aa_init(X, K = K, method = "random")
```

Uniform sampling is the simplest possible initialization and requires no
distance computations. Its expected quality depends entirely on the data
geometry. By design it explores the whole geometry with equal probability,
so it makes few assumptions about where useful archetypes should lie. When
combined with many random restarts via the `nrep` argument in `run_aa()`, even
simple uniform initialization can find good solutions as pointed out in
[@Krohne2019].

### `dirichlet`

Instead of sampling archetypes from the data, `dirichlet` constructs the
coefficients matrix $B$ directly by sampling each row from a symmetric Dirichlet
distribution $\text{Dirichlet}(\alpha \mathbf{1}_N)$. Thus the resulting
archetypes are random convex combinations of **all** data points.

The `alpha` parameter controls the shape of the distribution:

* **`alpha = 1`** (default): all possible combinations are equally likely
* **`alpha < 1`**: sparse combinations are more likely, so only a few points
  contribute to each archetype.
* **`alpha > 1`**: near uniform combinations are more likely, so many points
  contribute to each archetype.

As $\alpha \to 0$ the behavior approaches `random` (archetypes snap to data
points), while for $\alpha \gg 1$ the archetypes are more likely to lie near the
center of the data hull. The plot below compares initial archetype placements
across three variants to demonstrate the convergence of `dirichlet` to
`random`.

```{r dirichlet-vs-random-plot, fig.height = 3, fig.width=7, results='hold', echo=FALSE}
X <- as.matrix(toy)
K <- 3
dir_variants <- list(
    `random` = list(method = "random"),
    `dirichlet α = 1` = list(method = "dirichlet", alpha = 1),
    `dirichlet α = 0.01` = list(method = "dirichlet", alpha = 0.01)
)
n_dir_runs <- 3

dir_inits <- list()
for (variant in names(dir_variants)) {
    args <- dir_variants[[variant]]
    dir_inits[[variant]] <- list()
    for (run in 1:n_dir_runs) {
        if (args$method == "dirichlet") {
            dir_inits[[variant]][[run]] <- aa_init(
                X,
                K = K,
                method = args$method,
                alpha = args$alpha
            )
        } else {
            dir_inits[[variant]][[run]] <- aa_init(
                X,
                K = K,
                method = args$method
            )
        }
    }
}
op <- par(mfrow = c(1, 3), mar = c(2, 2, 2.5, 0.5), bty = "n")
run_cols <- adjustcolor(c("firebrick3", "darkorange3", "purple4"), 0.7)

for (nm in names(dir_inits)) {
    plot(
        X,
        pch = 16, cex = 0.45, col = "grey75",
        main = nm, xlab = "", ylab = "",
        xlim = range(X[, 1]) + c(-1, 1),
        ylim = range(X[, 2]) + c(-1, 1),
        axes = FALSE
    )
    for (run_ix in 1:length(dir_inits[[nm]])) {
        draw_simplex(
            dir_inits[[nm]][[run_ix]]$A,
            border = run_cols[run_ix],
            fill   = adjustcolor(run_cols[run_ix], 0.06),
            lty    = run_ix,
            cex    = 1.1
        )
    }
    if (nm == names(dir_inits)[1]) {
        legend(
            "topleft",
            legend = paste("Run", 1:n_dir_runs),
            col = run_cols, lty = 1:n_dir_runs, lwd = 2, pch = 17,
            bty = "n", cex = 0.7
        )
    }
}
par(op)
```

`"random"` archetypes (left) always coincide with data points and cluster
near high-density regions. `"dirichlet"` with $\alpha = 1$ (center) samples
weights from all data points and often produces archetypes near the center of
the hull. Lowering $\alpha$ (right) makes weights sparse, which pushes the
random mixtures toward specific data points and toward the behavior of
`"random"`.

`dirichlet` is used as the default option for the directional method because
directional data are represented through coefficients, so random convex
combinations are a natural starting point.

### Furthest First (`"furthest_first"`)

Selects the first archetype uniformly at random (biased toward points far from
the mean), then iteratively adds the single point that is *farthest* from the
already-selected set (the nearest-archetype distance to the set is maximized).

```{r furthest-first-ex, eval = FALSE}
X <- as.matrix(toy)
K <- 3
init <- aa_init(X, K = K, method = "furthest_first")
```

Furthest First is a greedy adaptation of the $k$-center algorithm. It produces
well-separated archetypes and is deterministic after the first random draw.
Because it targets the globally farthest point at each step, it tends to spread
archetypes evenly across the data hull, though this spread is not specifically
tuned to the AA reconstruction objective.

### k-means++ (`"kmeans_pp"`)

Selects the first archetype as in FurthestFirst, then samples each subsequent
archetype with probability proportional to the *squared* distance to the nearest
already-selected archetype.

```{r kmeans-pp-ex, eval = FALSE}
X <- as.matrix(toy)
K <- 3
init <- aa_init(X, K = K, method = "kmeans_pp")
```

The k-means++ sampling rule approximates AA++ (see below) by replacing the
convex-hull distance with the nearest-archetype distance [@Mair2023]. This makes
it substantially faster than exact AA++ while remaining competitive. It is a
good default when runtime is a concern but uniform sampling feels too
unpredictable.

### Furthest Sum (`"furthest_sum"`)

Selects archetypes by maximizing the *sum* of distances to all current
archetypes rather than the minimum [@Morup2012]. An optional refinement phase
cycles through the selected set and replaces each archetype with a better
candidate, controlled by `refinement_steps` (default 10).

```{r furthest-sum-ex, eval = FALSE}
X <- as.matrix(toy)
K <- 3
# Default refinement
init <- aa_init(X, K = K, method = "furthest_sum")

# More aggressive refinement
init <- aa_init(X, K = K, method = "furthest_sum", refinement_steps = 30)

# No refinement
init <- aa_init(X, K = K, method = "furthest_sum", refinement_steps = 0)
```

FurthestSum is the historical default initialization for AA [@Morup2012] and
remains the default when `init = NULL` is passed to `run_aa()`. It places
archetypes near the boundary of the data cloud and is computationally efficient.
See the [Initialization caveats](#initialization-caveats) below for an important
caveat about its interaction with standardization.

### Batched Coreset-Style Initialization

Supplying `batch_size` restricts initialization to a candidate batch before
selecting archetypes. With the default `batch_type = "distal"`, candidates are
sampled with probability proportional to each point's squared distance from the
data mean. This is the coreset idea of @Mair2019: reduce the effective problem
size while preserving the extremal structure that boundary-seeking methods such
as `furthest_sum` rely on.

```{r coreset-ex, eval = FALSE}
X <- as.matrix(toy)
K <- 3
batch_size <- 60

# batch_size must be at least K
init <- aa_init(X, K = K, method = "furthest_sum", batch_size = batch_size)
```

For large datasets where computing all pairwise distances is expensive, the
batch reduces memory use and runtime. Because distal batching downweights the
center, it avoids spending most candidate slots in regions where archetypes are
unlikely to be useful.

### AA++ (`"aa_pp"`)

AA++ [@Mair2023] selects each archetype by sampling with probability
proportional to the *squared residual distance from the convex hull* of the
already-selected archetypes. Concretely, after selecting $k - 1$ archetypes it
solves a small NNLS projection to find, for each data point, its best
approximation as a convex combination of the current archetypes; the squared
reconstruction error becomes the sampling weight for the $k$-th archetype.

```{r aa-pp-ex, eval = FALSE}
X <- as.matrix(toy)
K <- 3
init <- aa_init(X, K = K, method = "aa_pp")
```

Because the sampling probability is tied to the AA objective itself rather than
to a surrogate (such as nearest-archetype distance), AA++ has a formal guarantee:
each new archetype decreases the expected objective function (Proposition 3.2 of
@Mair2023). In broad empirical evaluations across many datasets, AA++
"almost always outperforms all baselines" [@Mair2023]. When $K = 2$, the convex
hull is a line segment and AA++ reduces to k-means++.

The main cost is one NNLS projection per archetype after the second, so
initializing $K$ archetypes requires $K - 2$ small optimization problems.
Despite this, initialization is still much cheaper than running $K$ full AA
iterations [@Mair2023].

### Batched AA++

AA++ can also use `batch_size`, giving a variant of the Monte Carlo AA++
approximation. Instead of projecting the full dataset at each step, it draws a
candidate batch, projects only those rows, and samples the next archetype from
their residuals. This is not identical to the `"aa_pp_mc"` scheme: candidate
batches follow `batch_type` and `batch_replace`, and the same batching interface
is used for the early AA++ selections. The `batch_size` parameter trades
approximation accuracy for speed and must satisfy `batch_size >= K`.

```{r aa-pp-mc-ex, eval = FALSE}
X <- as.matrix(toy)
K <- 3
batch_size <- 100
batch_type <- "uniform"

# Larger batch_size -> closer approximation to exact AA++
init <- aa_init(X,
    K = K,
    method = "aa_pp",
    batch_size = batch_size,
    batch_type = batch_type
)
```

When `batch_size == nrow(X)` the method reduces to exact AA++. When
`batch_size << nrow(X)` only a small fraction of the data is evaluated at each
step, making batched AA++ useful for large datasets where exact AA++ becomes
memory-bound. Use `batch_type = "uniform"` for the least biased approximation,
or keep the default `batch_type = "distal"` if you want the candidate batches
to emphasize likely boundary points.

### Hull-outmost (`"hull_outmost"`)

Builds a pool of candidate hull vertices, then ranks them by an outmost-vote
criterion: each candidate votes for the candidate farthest from it, and the $K$
most-voted candidates are selected as archetypes. The `hull_method` argument
(required) controls how the candidate pool is constructed:

* **`"full"`** — exact convex hull of $X$ in the original feature space.
  Uses `grDevices::chull` for 2-D data and the `geometry` package for higher
  dimensions.
* **`"projected"`** — union of convex hulls computed across random
  low-dimensional projections of $X$. Controlled by `projected_dim`
  (default `2`) and `n_projection_max`.
* **`"partitioned"`** — union of hull vertices from random partitions of the
  data rows. Controlled by `n_partitions` (default `10`). The fastest strategy.

```{r hull-outmost-ex, eval = FALSE}
X <- as.matrix(toy)
K <- 3
# Full hull — exact but requires 'geometry' for D > 2
init <- aa_init(X, K = K, method = "hull_outmost", hull_method = "full")

# Projected hull — works in any dimension without extra packages
init <- aa_init(X,
    K = K, method = "hull_outmost",
    hull_method = "projected", projected_dim = 2
)

# Partitioned hull — fastest, suitable for large n
init <- aa_init(X,
    K = K, method = "hull_outmost",
    hull_method = "partitioned", n_partitions = 15
)
```

The partitioned and projected strategies are designed for datasets where
computing the exact hull is intractable. The `use_unique_candidates` argument
(default `FALSE`) controls whether duplicate hull candidates (a point appearing
in multiple projections or partitions) cast only one vote.

## Session Information

```{r session-info, echo = FALSE}
sessionInfo()
```

## References
