---
title: "Choosing an Initialization Method for Archetypal Analysis"
bibliography: ../references.bib
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Choosing an Initialization Method for Archetypal Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5.5,
    fig.align = "center",
    warning = FALSE
)
set.seed(42)
par(bty = "n")
viridis_pal <- c("#440154", "#452B79", "#3A518A", "#2C718E", "#218F8B", "#28AD7F", "#5DC861", "#AADA31", "#FDE725")
library(YAAAP)
```

## Background

> This vignette assumes familiarity with the basic AA workflow covered in the
  **Introduction to Archetypal Analysis with YAAAP** vignette.

Archetypal analysis (AA) minimizes a reconstruction loss, written in the
standard Euclidean formulation as

$$\mathcal{L}(X, SA) = \|X - S A\|_F^2.$$

As with all the matrix decomposition problems, the objective is non-convex in
both $S$ and $A$, and gradient descent can terminate at a **local minimum**. As
a result, the quality of the final solution depends substantially on the starting
point.

YAAAP implements several options for initializing the archetype coordinates,
which have been suggested in the literature, through the `aa_init()` function
and the `init` argument.

## Methods at a Glance

The table below summarizes the eight methods by their strategy, whether they
involve random sampling, and any mandatory extra arguments.

| Method                 | Strategy                                                                           | Stochastic?      | Mandatory extras |
| ---------------------- | ---------------------------------------------------------------------------------- | ---------------- | :--------------: |
| `"random"`             | Sample $K$ rows uniformly at random                                                | Yes              | —                |
| `"dirichlet"`          | Draw archetypes as Dirichlet($\alpha$,...,$\alpha$) convex combinations            | Yes              | —                |
| `"kmeans_pp"`          | Sample proportional to squared distance to the nearest current archetype           | Yes              | —                |
| `"aa_pp"`              | Sample proportional to squared residual from the convex hull of current archetypes | Yes              | —                |
| `"aa_pp_mc"`           | MCMC approximation of AA++ using a Metropolis-Hastings batch                       | Yes              | `batch_size`     |
| `"furthest_first"`     | Always add the single point furthest from all current archetypes                   | Seed only        | —                |
| `"furthest_sum"`       | Maximize the *sum* of distances to all current archetypes; optional refinement     | Seed only        | —                |
| `"coreset_initfn"`     | Distance-weighted coreset subsample, then `furthest_sum` on the coreset            | Yes              | `m`              |
| `"hull_outmost"`       | Rank convex-hull candidates by outmost-vote frequency                              | Partitioned only | `hull_method`    |

## The `aa_init()` Function

`aa_init()` runs any initialization method as a standalone step, returning a
named list with two elements:

* **`A`** — $K \times M$ matrix of initial archetype coordinates.
* **`B`** — $K \times N$ row-stochastic matrix expressing each archetype as a
  convex combination of data points ($A = B X$).

```{r aa-init-basic}
toy <- read.csv(system.file("extdata", "toy.csv", package = "YAAAP"))
X <- as.matrix(toy) # 250 × 2

set.seed(42)
init <- aa_init(X, K = 3)
str(init)
```

The `A` matrix provides the initial archetype coordinates that will seed the
optimization loop; `B` is the convex-hull certificate for those coordinates. If
`run_aa()` is initialized with `aa_init()` and a method name or a custom function,
the `B` returned is carried into the algorithm initialization. If you instead
pass a pre-computed coordinate matrix such as `init$A`, `run_aa()` checks that
the coordinates still admit a feasible `B`. Rows outside the allowed data hull
are projected back into the hull with a warning.

```{r initialization-helpers, include = FALSE}
K <- 3
fit_tol <- 0.01
lbl <- c(
    random             = "Random",
    dirichlet          = "Dirichlet",
    kmeans_pp          = "k-means++",
    aa_pp              = "AA++",
    aa_pp_mc           = "AA++MC",
    furthest_first     = "Furthest First",
    furthest_sum       = "Furthest Sum",
    coreset_initfn     = "Coreset",
    hull_outmost       = "Hull-outmost"
)

method_args <- list(
    random             = list(),
    dirichlet          = list(),
    kmeans_pp          = list(),
    aa_pp              = list(),
    aa_pp_mc           = list(batch_size = 60),
    furthest_first     = list(),
    furthest_sum       = list(),
    coreset_initfn     = list(m = 60),
    hull_outmost       = list(hull_method = "partitioned")
)

make_init <- function(X, K, method, seed, extra = method_args[[method]]) {
    if (!missing(seed)) set.seed(seed)
    do.call(aa_init, c(list(X = X, K = K, method = method), extra))
}

draw_simplex <- function(A, border, fill = NULL, lty = 1, lwd = 2,
                         pch = 17, cex = 1.1) {
    if (nrow(A) >= 3) {
        hull <- chull(A)
        polygon(A[hull, , drop = FALSE],
            border = border, lwd = lwd, lty = lty,
            col = fill
        )
    } else if (nrow(A) == 2) {
        lines(A, col = border, lwd = lwd, lty = lty)
    }
    points(A, pch = pch, cex = cex, col = border)
}

nearest_rows <- function(X, A) apply(.pdist2(X, A), 2, which.min)

selection_deviation_score <- function(counts, K, n, pseudocount = 0.5) {
    total_counts <- sum(counts)
    N_runs <- total_counts / K
    expected <- total_counts / n
    p_null <- K / n

    log_fc <- log2((counts + pseudocount) / (expected + pseudocount))
    p_lower <- pbinom(counts, size = N_runs, prob = p_null)
    p_upper <- pbinom(counts - 1L, size = N_runs, prob = p_null,
        lower.tail = FALSE
    )
    p_value <- pmin(1, 2 * pmin(p_lower, p_upper))
    signed_log_p <- sign(counts - expected) *
        -log10(pmax(p_value, .Machine$double.xmin))

    list(log_fc = log_fc, p_value = p_value, signed_log_p = signed_log_p)
}

plot_selected_density <- function(X, counts, main, K, color_max = NULL,
                                  pseudocount = 0.5) {
    score <- selection_deviation_score(counts, K, nrow(X), pseudocount)
    signed_log_p <- score$signed_log_p

    pal <- colorRamp(c("blue", "white", "red"))
    cmax <- color_max
    if (is.null(cmax)) cmax <- max(abs(signed_log_p))
    probs <- if (cmax > 0) {
        pmax(0, pmin(1, (signed_log_p + cmax) / (2 * cmax)))
    } else {
        rep(0.5, length(signed_log_p))
    }
    point_cols <- pal(probs) / 255
    colnames(point_cols) <- c("red", "green", "blue")
    point_cols <- do.call(rgb, as.data.frame(point_cols))
    ix <- order(abs(signed_log_p))
    plot(
        X[ix, ],
        pch = 21, cex = 1, col = "gray50", bg = point_cols[ix],
        main = main, xlab = "", ylab = "", axes = FALSE, asp = 1
    )
}

plot_data_with_points <- function(X, A, main, col = "grey75", view = c(1, 2),
                                  pch = 4, point_col = "black") {
    plot(
        X[, view, drop = FALSE],
        pch = 16, cex = 0.45, col = col,
        main = main, xlab = "", ylab = "", axes = FALSE, asp = 1
    )
    points(A[, view, drop = FALSE],
        pch = pch, cex = 1.25, lwd = 1.6,
        col = point_col
    )
}

fit_from_method <- function(X, K, method, seed, nrep = 1, extra = method_args[[method]]) {
    if (!missing(seed)) set.seed(seed)
    run_aa(X,
        K = K, init = method, init_args = extra, nrep = nrep,
        scale = FALSE, max_iter = 60, tol = fit_tol
    )
}

# Make artificial datasets

make_concentric_circles <- function(n = 180, noise = 0.03) {
    n_inner <- n %/% 2
    n_outer <- n - n_inner
    theta   <- c(runif(n_inner, 0, 2 * pi),
                 runif(n_outer, 0, 2 * pi))
    radius  <- c(rep(0.4, n_inner),
                 rep(1.0, n_outer))
    X <- cbind(radius * cos(theta),
               radius * sin(theta))
    X <- X + rnorm(length(X), sd = noise)
    colnames(X) <- c("x", "y")
    list(
        X = X,
        view = c(1, 2),
        col = c(rep("#0072B2", n_inner), rep("#D55E00", n_outer)),
        main = "Concentric circles",
        kernel = "laplace",
        kernel_args = list(sigma = 0.25)
    )
}

make_swiss_roll <- function(n = 180, noise = 0.025) {
    t <- runif(n, 1.5 * pi, 4.5 * pi)
    h <- runif(n, -1, 1)
    X <- cbind(t * cos(t) / 7, h, t * sin(t) / 7)
    X <- X + rnorm(length(X), sd = noise)
    colnames(X) <- c("x", "height", "z")
    roll_cols <- colorRampPalette(c("#0072B2", "#009E73", "#D55E00"))(100)
    col_id <- pmax(1, pmin(100, ceiling(99 * (t - min(t)) / diff(range(t))) + 1))
    list(
        X = X,
        view = c(1, 3),
        col = roll_cols[col_id],
        main = "Swiss roll",
        kernel = "rbf",
        kernel_args = list(sigma = 0.45)
    )
}
```

## Supplying Your Own Initialization

The built-in method names cover the common cases, but `run_aa()` also accepts
either a precomputed matrix of archetype coordinates or a custom initialization
function.

If the starting coordinates are already available, pass them directly as a
matrix in the same units as the `X`. `run_aa()` will preprocess it the same way
it preprocesses `X`, handles the required checks and construct `B` construction.

```{r precomputed-init-matrix}
A0 <- X[seq_len(3), , drop = FALSE]

fit_matrix <- run_aa(X, K = 3, init = A0, scale = FALSE, tol = fit_tol)
```

A custom function is useful when initialization itself needs code. The function
must take `X` and `K`, and return the same two objects as `aa_init()`: the
archetype coordinates `A` and the data weights `B`. Additional arguments can
be passed through `init_args`. Below,is a deliberately simple initializer that
chooses the first `K` rows of the data.

```{r custom-init-function}
first_k_init <- function(X, K, ...) {
    A <- X[1:K, , drop = FALSE]
    B <- diag(1, nrow = K, ncol = nrow(X))
    list(A = A, B = B)
}

fit_custom <- run_aa(X, K = 3, init = first_k_init, scale = FALSE, tol = fit_tol)
```

If your custom initializer naturally produces only archetype coordinates,
`fit_simplex()` exposes the same simplex mapping used internally to construct a
feasible `B`. Reconstructing `A_init` as `B %*% X` ensures the coordinates passed
to AA are on the data hull:

```{r custom-init-fit-simplex, eval = FALSE}
A_user <- some_initializer(X, K = 3)
B <- fit_simplex(X, A_user)
A_init <- B %*% X
init <- list(A = A_init, B = B)
```

These two forms are useful when the starting point comes from a previous
analysis, a domain-specific rule, or an external clustering workflow. For most
workflows, passing a method name remains the shortest form.

```{r run-aa-init-short, eval = FALSE}
fit <- run_aa(X, K = 3, init = "furthest_sum", tol = 0.01)
fit <- run_aa(X, K = 3, init = "aa_pp", tol = 0.01)
fit <- run_aa(X,
    K = 3, init = "aa_pp_mc",
    init_args = list(batch_size = 200),
    tol = 0.01
)
```

## Comparing Initializer Biases

The first practical difference between initializers is not whether they are
"good" or "bad"; it is where they tend to look. The toy data below are simple:
250 two-dimensional points spread near the vertices of a triangle. We run all
built-in row-selecting initializers multiple times and count how many times
each data point is selected as an archetype.

```{r init-selection-density, cache=TRUE}
# Keep methods that select points
density_methods <- setdiff(names(method_args), "dirichlet")
N_expected <- 10  # Expected selections under random sampling
N_runs <- nrow(X) * N_expected

selected_counts <- lapply(density_methods, function(method) {
    counts <- integer(nrow(X))
    for (seed in seq_len(N_runs)) {
        init_obj <- make_init(X, K = K, method = method, seed = 999 + seed)
        selected <- which(init_obj$B > 0, arr.ind = TRUE)[, 2]
        counts[selected] <- counts[selected] + 1
    }
    counts
})
names(selected_counts) <- density_methods
```

```{r init-selection-density-plot, fig.width = 9, fig.height = 6.5, out.width="100%", echo = FALSE}
selection_color_max <- -log10(1 / nrow(X))

op <- par(mfrow = c(2, 4), mar = c(2, 2, 3, 0.5), bty = "n")
for (method in density_methods) {
    plot_selected_density(
        X,
        selected_counts[[method]],
        main = sprintf("%s\nselection deviation", lbl[method]),
        K = K,
        color_max = selection_color_max
    )
}
par(op)
```

As expected, `random` selects archetypes uniformly from the observed rows, so
the points do not diverge much from their null probability of selection. This
is probably not the best strategy for a convex geometry like the toy data, but
it is unbiased and thus robust to unknown data structure. With enough restarts
this strategy will eventually explore every region, but any single run may miss
the corners.

On the other extreme, methods like `furthest_first`, `furthest_sum`,
`coreset_initfn`, and `hull_outmost` are exploring more aggressively the
boundary of the data geometry. We see that most of them converge to the same set
of candidate points and never pick any point in the middle of of the cloud.
`furthest_first` is the only one that shows some variation in the selected
points, but it still has a strong corner preference. For a simple convex shape
that bias is often useful: the corners are exactly where good archetypes should
start.

Finally, in the middle between the two extremes, are clustering-inspired methods
like: `kmeans_pp`, `aa_pp`, and `aa_pp_mc`. These methods adapt ideas developed
to initialize clustering algorithms to the AA setting. They are more probabilistic
in nature and thus represent softer versions of their boundary searching counterparts.
They also rarely select any point in the center of the cloud but they are exploring
more diffusely the boundary regions and are not actively looking for the exterema.

More details about the methodology imployed by each method can be found in the
[Method Reference](#method-reference) section at the end of this vignette.
The rest of the vignette narrows to three representatives: `random` as an
exploratory baseline, `kmeans_pp` as the softer outward-biased initializer, and
`furthest_sum` for the strong greedy boundary bias.

## Increasing the Number of Archetypes

With $K = 3$, the toy data roughly match the three-corner shape. When $K$
increases, the question changes from "did we find the corners?" to "how do the
extra archetypes cover the cloud?" The following chunk is the user-facing call:
only the `method` and `K` values change.

```{r increasing-k-user-code}
set.seed(42)
init_k7 <- aa_init(X, K = 7, method = "kmeans_pp")
```

```{r increasing-k-setup, include = FALSE}
k_grid <- c(3, 7, 15)
k_methods <- c("random", "kmeans_pp", "furthest_sum")
k_inits <- lapply(k_methods, function(method) {
    lapply(k_grid, function(k_value) {
        make_init(X, K = k_value, method = method, seed = 1600 + k_value)
    })
})
names(k_inits) <- k_methods
k_inits <- lapply(k_inits, setNames, paste0("K = ", k_grid))
```

```{r increasing-k-plot, fig.width = 9, fig.height = 6.5, out.width="100%", echo = FALSE}
op <- par(
    mfrow = c(length(k_methods), length(k_grid)),
    mar = c(0.5, 0.5, 2.2, 0.2), oma = c(0, 0, 0, 0), bty = "n"
)
for (method in k_methods) {
    for (k_label in names(k_inits[[method]])) {
        init_obj <- k_inits[[method]][[k_label]]
        plot(
            X,
            pch = 16, cex = 0.42, col = "grey75",
            main = sprintf("%s\n%s", lbl[method], k_label),
            xlab = "", ylab = "", axes = FALSE, asp = 1
        )
        draw_simplex(
            init_obj$A,
            border = "firebrick3",
            fill = adjustcolor("firebrick3", 0.05),
            cex = 0.95
        )
    }
}
par(op)
```

As $K$ grows, the non-aggressive methods start to waste extra archetypes in
the center of the convex hull. This is visible for `random`, and eventually also
for `kmeans_pp`: once the main corners have been sampled, additional points are
often placed in the interior regions where there are definitely no archetypes
for this triangle. Those starts can still optimize successfully, but the
initialization budget is no longer being used to approximate the boundary.

`furthest_sum` behaves differently. It keeps adding points on the outside of the
cloud and, as $K$ increases, the selected polygon becomes a finer approximation
to the true convex hull boundary. In fact, `furthest_sum` generated points
are "guaranteed to lie in the minimal convex set of the unselected data points"
[Theorem 2 in [@Morup2012]]

## When Initialization Matters

The toy triangle is intentionally simple. If we start from one initialization
and then let `run_aa()` optimize, most methods reach essentially the same final
fit. The red dashed triangle below is the starting point; the blue solid
triangle is the fitted solution.

```{r init-before-after-fit}
focused_methods <- c("random", "kmeans_pp", "furthest_sum")
focused_inits <- lapply(seq_along(focused_methods), function(i) {
    method <- focused_methods[i]
    make_init(X, K = K, method = method, seed = 2000 + i)
})
names(focused_inits) <- focused_methods

fits_from_init <- lapply(focused_inits, function(init_obj) {
    run_aa(X, K = K, init = init_obj$A, scale = FALSE, max_iter = 60,
        tol = fit_tol
    )
})
fit_steps <- vapply(
    fits_from_init, function(fit) as.integer(nrow(fit$loss) - 1),
    integer(1)
)
```

```{r init-before-after-plot, fig.width = 8.5, fig.height = 6, out.width="100%", echo = FALSE}
loss_ylim <- range(unlist(lapply(fits_from_init, function(fit) fit$loss$loss)))
op <- par(mfrow = c(2, 3), mar = c(2.2, 2.5, 3, 0.5), bty = "n")
for (method in focused_methods) {
    plot(
        X,
        pch = 16, cex = 0.45, col = "grey75",
        main = sprintf("%s\n%d steps", lbl[method], fit_steps[method]),
        xlab = "", ylab = "", axes = FALSE, asp = 1
    )
    draw_simplex(
        focused_inits[[method]]$A,
        border = adjustcolor("firebrick3", 0.75),
        fill = adjustcolor("firebrick3", 0.05),
        lty = 2
    )
    draw_simplex(
        fits_from_init[[method]]$coordinates,
        border = "steelblue4",
        fill = adjustcolor("steelblue4", 0.08),
        pch = 19
    )
    if (method == focused_methods[1]) {
        legend(
            "topleft",
            legend = c("Initial", "Fitted"),
            col = c("firebrick3", "steelblue4"), lty = c(2, 1), lwd = 2,
            pch = c(17, 19), bty = "n", cex = 0.7
        )
    }
}
for (method in focused_methods) {
    loss <- fits_from_init[[method]]$loss
    iterations <- seq_len(nrow(loss)) - 1L
    plot(
        iterations, loss$loss, type = "l", lwd = 1.8, col = "steelblue4",
        ylim = loss_ylim, main = "Loss", xlab = "Iteration", ylab = "",
        las = 1
    )
    points(iterations, loss$loss, pch = 16, cex = 0.45, col = "steelblue4")
}
par(op)
```

This is the reassuring case: the data geometry is simple enough that a mediocre
start can still be corrected by the optimizer. Initialization matters more when
the geometry is less well described by straight-line Euclidean distances.

## Nonlinear Geometries

The previous examples still live in a simple Euclidean geometry: distances in
the plotted coordinates describe the structure we care about. Some datasets are
not like that. In concentric circles, the inner circle is important even though
the outer circle contains the most distant points. In a Swiss roll, points that
look close in a two-dimensional projection may be far apart along the rolled
surface.

```{r nonlinear-data, include = FALSE}
set.seed(3001)
nonlinear_data <- list(
    circles = make_concentric_circles(),
    swiss_roll = make_swiss_roll()
)
nonlinear_K <- 8
nonlinear_methods <- c("random", "kmeans_pp", "furthest_sum")
nonlinear_inits <- lapply(nonlinear_data, function(data_obj) {
    out <- lapply(nonlinear_methods, function(method) {
        make_init(
            data_obj$X,
            K = nonlinear_K,
            method = method,
            seed = 3010 + match(method, nonlinear_methods)
        )
    })
    names(out) <- nonlinear_methods
    out
})
```

First inspect the Euclidean initializations. The biased methods are doing what
they were designed to do: they seek far-away points. On these shapes, that can
over-emphasize the outside and leave the center or inner structure
under-represented. `random` does not fix the geometry, but it is less committed
to the outside by construction.

```{r nonlinear-init-user-code}
set.seed(3011)
circle_init <- aa_init(nonlinear_data$circles$X,
    K = 8,
    method = "furthest_sum"
)
```

```{r nonlinear-init-plot, fig.width = 8.5, fig.height = 6, out.width="100%", echo = FALSE}
op <- par(
    mfrow = c(length(nonlinear_data), length(nonlinear_methods)),
    mar = c(2, 2, 3, 0.5), bty = "n"
)
for (dataset_name in names(nonlinear_data)) {
    data_obj <- nonlinear_data[[dataset_name]]
    for (method in nonlinear_methods) {
        init_obj <- nonlinear_inits[[dataset_name]][[method]]
        plot_data_with_points(
            data_obj$X,
            init_obj$A,
            main = sprintf("%s\n%s", data_obj$main, lbl[method]),
            col = adjustcolor(data_obj$col, 0.62),
            view = data_obj$view
        )
    }
}
par(op)
```

For these cases, the better answer is often to change the space in which AA is
fit. A kernel method lets nearby points influence each other according to a
smooth similarity rule rather than only through straight-line distance in the
input coordinates.

```{r kernel-user-code}
fit_kernel <- archetypes_kernel_pgd(
    nonlinear_data$circles$X,
    K = 8,
    kernel = "laplace",
    kernel_args = list(sigma = 0.25),
    init = "furthest_sum",
    tol = fit_tol
)
```

```{r nonlinear-kernel-setup, include = FALSE}
kernel_methods <- c("furthest_sum", "kmeans_pp")
kernel_inits <- lapply(names(nonlinear_data), function(dataset_name) {
    data_obj <- nonlinear_data[[dataset_name]]
    out <- lapply(kernel_methods, function(method) {
        euclidean_A <- nonlinear_inits[[dataset_name]][[method]]$A

        set.seed(4100 + match(method, kernel_methods))
        kernel_fit <- suppressWarnings(archetypes_kernel_pgd(
            data_obj$X,
            K = nonlinear_K,
            kernel = data_obj$kernel,
            kernel_args = data_obj$kernel_args,
            init = method,
            max_iter = 1,
            tol = fit_tol,
            tol_r2 = 1
        ))
        kernel_A <- kernel_fit$init %*% data_obj$X
        list(euclidean = euclidean_A, kernel = kernel_A)
    })
    names(out) <- kernel_methods
    out
})
names(kernel_inits) <- names(nonlinear_data)
```

```{r nonlinear-kernel-plot, fig.width = 8.5, fig.height = 6, out.width="100%", echo = FALSE}
op <- par(
    mfrow = c(length(nonlinear_data), length(kernel_methods)),
    mar = c(2, 2, 3, 0.5), bty = "n"
)
for (dataset_name in names(nonlinear_data)) {
    data_obj <- nonlinear_data[[dataset_name]]
    for (method in kernel_methods) {
        init_obj <- kernel_inits[[dataset_name]][[method]]
        plot(
            data_obj$X[, data_obj$view, drop = FALSE],
            pch = 16, cex = 0.45, col = adjustcolor(data_obj$col, 0.55),
            main = sprintf("%s\n%s", data_obj$main, lbl[method]),
            xlab = "", ylab = "", axes = FALSE, asp = 1
        )
        points(init_obj$euclidean[, data_obj$view, drop = FALSE],
            pch = 4, cex = 1.2, lwd = 1.5, col = "black"
        )
        points(init_obj$kernel[, data_obj$view, drop = FALSE],
            pch = 1, cex = 1.25, lwd = 1.6, col = "#D55E00"
        )
        if (dataset_name == names(nonlinear_data)[1] &&
            method == kernel_methods[1]) {
            legend(
                "topright",
                legend = c("Euclidean init", "Kernel init"),
                pch = c(4, 1), col = c("black", "#D55E00"),
                pt.cex = c(1.1, 1.2), bty = "n", cex = 0.7
            )
        }
    }
}
par(op)
```

The same initialization names still work. The important difference is that the
kernel version measures spread in the similarity space used by the model. This
can rescue boundary-seeking methods on shapes where the raw coordinates make
the outside look more important than the rest of the structure.

## Practical Guidance

There is no universal best initializer. The right choice depends on how much
you know about the geometry.

| Situation                             | Recommended `init`                                                        |
| ------------------------------------- | ------------------------------------------------------------------------- |
| Simple convex-ish data                | `"furthest_sum"`, `"kmeans_pp"`, or `"aa_pp"`                            |
| Unknown structure, enough restarts    | `"random"` with a larger `nrep`                                           |
| Unknown structure, few restarts       | `"aa_pp"` or `"kmeans_pp"`                                                 |
| Clearly nonlinear geometry           | `method = "kernel"` with a suitable kernel                                |
| Large dataset ($n \gg 10{,}000$)      | `"aa_pp_mc"` with a moderate `batch_size`                                 |
| Very large dataset, speed critical    | `"coreset_initfn"` or `"hull_outmost"` with `hull_method = "partitioned"` |
| Reproducing legacy results            | `"furthest_sum"` (historical AA default, @Morup2012)                      |

When the geometry is simple, a boundary or coverage bias can be helpful. When
the geometry is unknown, `random` with enough restarts is often the safest
baseline because it makes the fewest assumptions. If the geometry is visibly
nonlinear, changing the model with `archetypes_kernel_pgd()` is usually more
important than fine-tuning the Euclidean initializer.

## Method Reference {#method-reference}

### `random`

Selects $K$ rows of $X$ uniformly at random without replacement.

```{r uniform-ex, eval = FALSE}
init <- aa_init(X, K = 3, method = "random")
```

Uniform sampling is the simplest possible initialization and requires no
distance computations. Its expected quality depends entirely on the data
geometry. By design it explores the whole geometry with equal probability,
so it is unaffected by potential biases in the data distribution trapping it
in a local minimum. When combined with many random restarts via the `nrep`
argument in `run_aa()`, even simple uniform initialization can find good
solutions as pointed out in [@Krohne2019].

### `dirichlet`

Instead of sampling archetypes from the data, `dirichlet` constructs the
coefficients matrix $B$ directly by sampling each row from a symmetric Dirichlet
distribution $\text{Dirichlet}(\alpha \mathbf{1}_N)$. Thus the resulting
archetypes are random convex combinations of **all** data points.

The `alpha` parameter controls the shape of the distribution:

* **`alpha = 1`** (default): all possible combinations are equally likely
* **`alpha < 1`**: sparse combinations are more likely thus only a few points
  contribute to each archetype.
* **`alpha > 1`**: near uniform combinations are more likely thus many points
  contribute to each archetype.

So as $\alpha \to 0$ the behavior approaches `random` (archetypes snap to data points),
while for $\alpha \gg 1$ the archetypes are more likely to lie near the center of
the data hull. The plot below compares initial archetype placements across three
variants to demonstrate the convergence of `dirichlet` to `random`.

```{r dirichlet-vs-random-setup}
dir_variants <- list(
    `random` = list(method = "random"),
    `dirichlet α = 1` = list(method = "dirichlet", alpha = 1),
    `dirichlet α = 0.01` = list(method = "dirichlet", alpha = 0.01)
)
n_dir_runs <- 3
dir_seeds <- 42 + seq_len(n_dir_runs) - 1

dir_inits <- lapply(dir_variants, function(args) {
    lapply(dir_seeds, function(s) {
        set.seed(s)
        do.call(aa_init, c(list(X = X, K = K), args))
    })
})
```

```{r dirichlet-vs-random-plot, echo = FALSE, fig.height = 3, fig.width=7}
op <- par(mfrow = c(1, 3), mar = c(2, 2, 2.5, 0.5), bty = "n")
run_cols <- adjustcolor(c("firebrick3", "darkorange3", "purple4"), 0.7)

for (nm in names(dir_inits)) {
    plot(
        X,
        pch = 16, cex = 0.45, col = "grey75",
        main = nm, xlab = "", ylab = "",
        xlim = range(X[, 1]) + c(-1, 1),
        ylim = range(X[, 2]) + c(-1, 1),
        axes = FALSE
    )
    for (run_ix in seq_along(dir_inits[[nm]])) {
        draw_simplex(
            dir_inits[[nm]][[run_ix]]$A,
            border = run_cols[run_ix],
            fill   = adjustcolor(run_cols[run_ix], 0.06),
            lty    = run_ix,
            cex    = 1.1
        )
    }
    if (nm == names(dir_inits)[1]) {
        legend(
            "topleft",
            legend = paste("Run", seq_len(n_dir_runs)),
            col = run_cols, lty = seq_len(n_dir_runs), lwd = 2, pch = 17,
            bty = "n", cex = 0.7
        )
    }
}
par(op)
```

`"random"` archetypes (left) always coincide with data points and cluster
near high-density regions. `"dirichlet"` with $\alpha = 1$ (center) select a
random distribution of weights from all data points and because there are more
of them near the "uniform" distribution tends to produce archetypes that lie
near the center of the hull. Lowering $\alpha$ (right) biases the results towards
distribution with a few large weights, which pushes the random mixtures towards
specific data points thus approaching the behavior of `"random"`.

`dirichlet` is used as the default option for the directional method because directional data are represented through coefficients, so random convex combinations are a natural starting point.

### Furthest First (`"furthest_first"`)

Selects the first archetype uniformly at random (biased toward points far from
the mean), then iteratively adds the single point that is *farthest* from the
already-selected set (the nearest-archetype distance to the set is maximized).

```{r furthest-first-ex, eval = FALSE}
init <- aa_init(X, K = 3, method = "furthest_first")
```

FurthestFirst is a greedy adaptation of the $k$-center algorithm.  It produces
well-separated archetypes and is deterministic after the first random draw.
Because it targets the globally farthest point at each step, it tends to spread
archetypes evenly across the data hull, though this spread is not specifically
tuned to the AA reconstruction objective.

### k-means++ (`"kmeans_pp"`)

Selects the first archetype as in FurthestFirst, then samples each subsequent
archetype with probability proportional to the *squared* distance to the nearest
already-selected archetype.

```{r kmeans-pp-ex, eval = FALSE}
init <- aa_init(X, K = 3, method = "kmeans_pp")
```

The k-means++ sampling rule approximates AA++ (see below) by replacing the
convex-hull distance with the nearest-archetype distance [@Mair2023]. This makes
it substantially faster than exact AA++ while remaining competitive. It is a
good default when runtime is a concern but uniform sampling feels too
unpredictable.

### Furthest Sum (`"furthest_sum"`)

Selects archetypes by maximizing the *sum* of distances to all current
archetypes rather than the minimum [@Morup2012]. An optional refinement phase
cycles through the selected set and replaces each archetype with a better
candidate, controlled by `refinement_steps` (default 10).

```{r furthest-sum-ex, eval = FALSE}
# Default refinement
init <- aa_init(X, K = 3, method = "furthest_sum")

# More aggressive refinement
init <- aa_init(X, K = 3, method = "furthest_sum", refinement_steps = 30)

# No refinement
init <- aa_init(X, K = 3, method = "furthest_sum", refinement_steps = 0)
```

FurthestSum is the historical default initialization for AA [@Morup2012] and
remains the default when `init = NULL` is passed to `run_aa()`. It places
archetypes near the boundary of the data cloud and is computationally efficient.
See the [Initialization caveats](#initialization-caveats) below for an important caveat
about its interaction with standardization.

### Coreset + FurthestSum (`"coreset_initfn"`)

Subsamples a coreset of $m$ points from $X$ with probability proportional to
each point's squared distance from the data mean, then runs `furthest_sum` on
the coreset [@Mair2019]. The mandatory `m` parameter controls the coreset size
and must satisfy `m >= K`.

```{r coreset-ex, eval = FALSE}
# m must be at least K
init <- aa_init(X, K = 3, method = "coreset_initfn", m = 60)
```

For large datasets where computing all pairwise distances is expensive, the
coreset strategy reduces the effective problem size. Because the coreset is
distance-weighted, the approximation preserves the extremal structure that
FurthestSum relies on [@Mair2019].

### AA++ (`"aa_pp"`)

AA++ [@Mair2023] selects each archetype by sampling with probability
proportional to the *squared residual distance from the convex hull* of the
already-selected archetypes. Concretely, after selecting $k - 1$ archetypes it
solves a small NNLS projection to find, for each data point, its best
approximation as a convex combination of the current archetypes; the squared
reconstruction error becomes the sampling weight for the $k$-th archetype.

```{r aa-pp-ex, eval = FALSE}
init <- aa_init(X, K = 3, method = "aa_pp")
```

Because the sampling probability is tied to the AA objective itself rather than
to a surrogate (such as nearest-archetype distance), AA++ has a formal guarantee:
each new archetype decreases the expected objective function (Proposition 3.2 of
@Mair2023). In broad empirical evaluations across many datasets, AA++
"almost always outperforms all baselines" [@Mair2023]. When $K = 2$, the convex
hull is a line segment and AA++ reduces to k-means++.

The main cost is one NNLS projection per archetype after the second, so
initializing $K$ archetypes requires $K - 2$ small optimization problems.
Despite this, initialization is still much cheaper than running $K$ full AA
iterations [@Mair2023].

### AA++MC (`"aa_pp_mc"`)

AA++MC approximates AA++ through a Metropolis–Hastings Markov chain [@Mair2023].
Instead of projecting the full dataset at each step, it draws a random batch of
`batch_size` points, projects them, and uses the Metropolis acceptance rule to
select the next archetype candidate. The mandatory `batch_size` parameter trades
approximation accuracy for speed and must satisfy `batch_size >= K`.

```{r aa-pp-mc-ex, eval = FALSE}
# Larger batch_size → closer approximation to exact AA++
init <- aa_init(X, K = 3, method = "aa_pp_mc", batch_size = 100)
```

When `batch_size == nrow(X)` the Metropolis acceptance probability is always 1
and the method reduces to exact AA++. When `batch_size << nrow(X)` only a
small fraction of the data is evaluated at each step, making AA++MC well-suited
for large datasets where exact AA++ becomes memory-bound.

### Hull-outmost (`"hull_outmost"`)

Builds a pool of candidate hull vertices, then ranks them by an outmost-vote
criterion: each candidate votes for the candidate farthest from it, and the $K$
most-voted candidates are selected as archetypes. The `hull_method` argument
(required) controls how the candidate pool is constructed:

* **`"full"`** — exact convex hull of $X$ in the original feature space.
  Uses `grDevices::chull` for 2-D data and the `geometry` package for higher
  dimensions.
* **`"projected"`** — union of convex hulls computed across random
  low-dimensional projections of $X$. Controlled by `projected_dim`
  (default `2`) and `n_projection_max`.
* **`"partitioned"`** — union of hull vertices from random partitions of the
  data rows. Controlled by `n_partitions` (default `10`). The fastest strategy.

```{r hull-outmost-ex, eval = FALSE}
# Full hull — exact but requires 'geometry' for D > 2
init <- aa_init(X, K = 3, method = "hull_outmost", hull_method = "full")

# Projected hull — works in any dimension without extra packages
init <- aa_init(X,
    K = 3, method = "hull_outmost",
    hull_method = "projected", projected_dim = 2
)

# Partitioned hull — fastest, suitable for large n
init <- aa_init(X,
    K = 3, method = "hull_outmost",
    hull_method = "partitioned", n_partitions = 15
)
```

The partitioned and projected strategies are designed for datasets where
computing the exact hull is intractable. The `use_unique_candidates` argument
(default `FALSE`) controls whether duplicate hull candidates (a point appearing
in multiple projections or partitions) cast only one vote.

## Initialization Caveats {#initialization-caveats}

The default `scale = TRUE` in `run_aa()` applies column-wise standardization
(z-scoring) before fitting. While scaling often improves the conditioning of the
optimization, it interacts differently with each initialization strategy.

@Mair2023 demonstrate that FurthestSum is "especially sensitive to the
pre-processing scheme": standardization can degrade its performance because it
distorts the Euclidean distances that FurthestSum relies on to place archetypes
near the data boundary. AA++ and its approximations are robust to this effect
because their sampling probabilities are tied to the actual objective loss
rather than to raw Euclidean distances [@Mair2023].

Additionally, @Suleman2017 reports that FurthestSum tends to produce redundant
(near-duplicate) archetypes when $K$ is large relative to the intrinsic
dimensionality of the data. If you observe that several fitted archetypes are
very similar to one another, switching to `"aa_pp"` is recommended.

Initializer names are not supported identically by every model family. The
Euclidean solvers (`method = "pgd"`, `"nnls"`, and `"paa"`) call
`aa_init()` on the preprocessed data, so they accept the full catalogue of
initializer strings described above, as well as custom initializer functions and
coordinate matrices.

Kernel AA is the main exception. With `method = "kernel"`, initializer strings
are limited to `"random"`, `"furthest_first"`, `"kmeans_pp"`, and
`"furthest_sum"`. These methods can be expressed using selected row indices and
pairwise distances from the kernel Gram matrix. The other strings are not
accepted as kernel initializers: `"dirichlet"` creates arbitrary coefficient
mixtures, `"aa_pp"` and `"aa_pp_mc"` require convex-hull residual projections,
`"coreset_initfn"` requires a coreset construction tied to the input-space
geometry, and `"hull_outmost"` requires explicit hull candidates. Those
operations would need separate kernel-space implementations. If you need one of
those starts for a kernel fit, pass an explicit `K x n` non-negative coefficient
matrix instead; row indices or row names are also accepted when you want to
select observed rows directly.

The directional solver also accepts the `aa_init()` method
strings, but it defaults to `"dirichlet"` for historical reasons^[the paper
introducing it opted for this method].

## Session Information

```{r session-info, echo = FALSE}
sessionInfo()
```

## References
