---
title: "Introduction to Archetypal Analysis with YAAAP"
bibliography: ../references.bib
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to Archetypal Analysis with YAAAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 5,
  fig.align = "center"
)
set.seed(42L)
par(bty = "n", frame.plot = FALSE)
```

## What is Archetypal Analysis?

Archetypal analysis (AA) is an unsupervised learning method that describes every
observation as a *convex mixture*^[weighted sums with non-negative weights
summing to 1] of a small number of extreme, idealized profiles called
**archetypes** [@Cutler1994]. Unlike PCA, whose components are abstract
mathematical directions, or k-means, whose clusters are represented by
centroids, archetypes are located at the *boundary* of the data representing the
distinct extremes. This gives AA an appealing balance between the interpretability
of clustering and the flexibility of matrix factorization. In practice,
archetypes often correspond to "pure" or "extreme" profiles, making the method
especially useful in settings where observations are naturally viewed as
mixtures of biologically, chemically, or behaviorally distinct states.

Formally, given a data matrix $X \in \mathbb{R}^{N \times M}$ (N samples, M
features), AA finds:

- **Archetypes** $A \in \mathbb{R}^{K \times M}$ : K extreme profiles, each of
  which is itself a weighted average of actual data points ($A = BX$, with
  $B \geq 0$, $\mathbf{1}^\top B = \mathbf{1}$),
- **Compositions** $S \in \mathbb{R}^{N \times K}$ : for each observation, a
  non-negative weight vector summing to 1 ($S \geq 0$,
  $S\mathbf{1} = \mathbf{1}$),

such that the reconstruction $\hat{X} = SA$ is as close as possible to $X$ in
the standard squared Frobenius norm^[sum of squared residuals]:

$$\min_{B,S} \|X - SBX\|_F^2 \quad \text{s.t.} \quad A = BX,\; B,S \geq 0,\;
  B\mathbf{1} = S\mathbf{1} = \mathbf{1}.$$

The constraint $A = BX$ keeps archetypes *inside* (or on the boundary of) the
convex hull of the data, making them directly interpretable as extreme
prototypes. The composition vectors $S$ correspond to points in the archetype
simplex: a composition of $(1,0,0)$ means a sample is entirely characterised by
archetype 1, while $(1/3,1/3,1/3)$ means it sits equally between all three.

### Relationship to other methods

|Method              |Main_focus                             |Sample_representation                     |Component_type               |Interpretability      |Flexibility      |
|:-------------------|:--------------------------------------|:-----------------------------------------|:----------------------------|:---------------------|:----------------|
|PCA / SVD           |Directions of maximal variance         |Linear combination                        |Abstract latent directions   |Often moderate to low |High             |
|NMF                 |Additive parts-based representation    |Non-negative combination                  |Non-negative latent parts    |Moderate to high      |Moderate to high |
|k-means             |Central cluster prototypes             |Hard assignment to one cluster            |Cluster centroids            |High                  |Low to moderate  |
|k-medoids           |Representative observed prototypes     |Hard assignment to one observed prototype |Observed data points         |High                  |Low              |
|Archetypal Analysis |Extreme profiles / corners of the data |Convex combination of archetypes          |Observed-data-based extremes |High                  |Moderate to high |

---

## The YAAAP package

**YAAAP** (Yet Another Archetypal Analysis Package) provides fast, flexible AA
for R. This vignette covers the essentials:

1. Fitting a model with `run_aa()`, from both a matrix and a formula.
2. Inspecting the result object of class `archetypes`.
3. Visualising archetypes and compositions.
4. Predicting compositions for new data.
5. Practical considerations: scaling, restarts, convergence, and the `delta` relaxation.
6. The algorithm and its key arguments.
7. Robust fitting and missing data.

```{r library}
library(YAAAP)
```

## A first example: the toy dataset

The package ships a small two-dimensional `toy` dataset (originally from the
[`archetypes`](https://cran.r-project.org/package=archetypes)) that contains
250 points scattered around three vertices of a triangle.

```{r load-toy}
toy <- read.csv(system.file("extdata", "toy.csv", package = "YAAAP"))
head(toy, 3)
dim(toy)
```

```{r plot-toy, fig.cap = "The toy dataset: three clusters near the corners of a triangle."}
plot(toy, asp = 1., frame.plot = FALSE, main = "Toy dataset",
     pch = 19, cex = 0.6, col = "grey50")
```

We fit an AA model with $K = 3$ archetypes using `run_aa()`.

```{r fit-toy}
fit <- run_aa(toy, K = 3L, nrep = 3L)
fit
```

The output reports the final $R^2$ and convergence status.

### The `archetypes` result object

`run_aa()` returns an S3 object of class `archetypes`.
It is a list with the following main components:

| Component      | Dimensions | Description                                             |
|----------------|------------|---------------------------------------------------------|
| `coordinates`  | K x M      | Archetype profiles in feature space                     |
| `coefficients` | K x N      | Weights expressing each archetype as a mix of samples   |
| `compositions` | N x K      | Barycentric coordinates of each sample                  |
| `loss`         | (iter) x 4 | Per-iteration RSS, $R^2$, and condition numbers         |
| `converged`    | scalar     | Whether the optimiser converged                         |

```{r inspect}
fit$coordinates
dim(fit$compositions)
```

The fundamental identity $\hat{X} = SA$ is:

```{r reconstruction, eval=FALSE}
X_hat <- fit$compositions %*% fit$coordinates
```

Let us verify this recovers the data well:

```{r identity-check}
X <- as.matrix(toy)
X_hat <- fitted(fit)
R <- resid(fit)  # residuals: X - X_hat
max(rowSums(R * R) / rowSums(X * X))  # max relative reconstruction error
norm(X - X_hat, "F")^2  # residual sum of squares
```

The `$loss` data frame records four diagnostics at each logged iteration:

| Column | Meaning                                                                              |
|--------|--------------------------------------------------------------------------------------|
| `loss` | Working objective (typically $\|X - SA\|_F^2$, evaluated on the internally scaled data) |
| `r2`   | $1 - \text{RSS}/\text{TSS}$ — proportion of variance explained                      |
| `k_S`  | Condition number of $S$ — high $\to$ near-degenerate compositions                   |
| `k_A`  | Condition number of $A$ — high $\to$ near-collinear archetypes                      |

Note that `loss` values may differ from the RSS computed directly above because
the working objective operates on internally scaled data. See <a href="#algorithm-details">*Algorithm details*</a>
for a full discussion.

### Visualising the fit

`plot()` method for `archetypes` objects offers several `what` options.

#### Archetype positions in feature space

This method is provided for convenience and is equivalent to plotting the
original data points and then overlaying the archetypes from the `coordinates` matrix
with lines connecting them.

```{r plot-coordinates, fig.cap = "Data and estimated archetypes. The triangle connects the three archetypes."}
plot(fit, what = "coordinates",
     main = "Archetype positions in feature space",
     asp = 1, frame.plot = FALSE,
     pch = 19, cex = 0.5, col = "grey50",
    # use args.archetypes to configure archetype aesthetics separately
     args.archetypes = list(pch = 17, cex = 1.5, col = "firebrick"))
```

As expected, the archetypes lie close to the corners of the point cloud.

## Working with "real" data: Fisher's iris

Besides tabular data (like `data.frame` or `matrix`), `run_aa()` also accepts a
**formula** and **data frame** combinations. Using a formula is the most convenient
when the data contains a response column you wish to exclude: the response is silently
dropped and all right-hand-side predictors are used as features.

```{r fit-iris}
# 'Species' (the response) is discarded; the four measurements are used.
set.seed(1L)
fit_iris <- run_aa(Species ~ ., data = iris, K = 3L, scale = TRUE, tol = 1e-4)
fit_iris
```

This is equivalent to calling `run_aa(iris[, 1:4], ...)`

Because the four measurements are on different scales the default
`scale = TRUE` (column-wise z-scoring) is appropriate.

In the following subsections, we build analyse the fitted model in a typical workflow:

- we first locate the archetypes in feature space (PCA projection),
- then identify them from the observations that define them and
- characterise what they represent through a feature-profile plot.
- We then explore how the 150 samples distribute across the archetype simplex, and
- finally show how the fitted model predicts compositions for new observations.

### Archetype positions in feature space

Since the iris dataset has four features, the coordinate scatter requires
dimensionality reduction. Pass `projection = "pca"` to project both data and
archetypes onto the first two principal components.

```{r iris-coordinates, fig.cap = "PCA projection of the iris data and its three archetypes (triangles)."}
pal_sp <- c(setosa = "#E69F00", versicolor = "#56B4E9", virginica = "#009E73")
plot(fit_iris, what = "coordinates", frame.plot = FALSE, asp = 1,
     projection = "pca",
     col = pal_sp[as.character(iris$Species)],
     pch = 19,
     cex = 0.6,
     args.archetypes = list(col = "firebrick", pch = 17, cex = 1.5))
```

### Naming archetypes from their defining observations

Each archetype is a convex mixture of actual data points, described by the
`coefficients` matrix (K x N). To give more meaningful names to the archetypes,
we can look at which species contribute most to each archetype and assign
names accordingly.

```{r iris-species-archetypes-match}
B <- coef(fit_iris)  # K x N

# Sum coefficients within each species (rows = species, cols = archetypes)
coef_by_species <- aggregate(t(B), by = iris["Species"], FUN = sum)
knitr::kable(
    coef_by_species,
    digits = 3,
    caption = "Species contribution to Archetypes"
)

```

From the table we see that one archetype represents a mixture of "setosa" and
"versicolor", while the other two are predominantly "setosa" and "virginica"
respectively. We label the archetypes accordingly.

```{r iris-rename-archetypes}
# Concat all species contributing to an archetype
coef_mat <- as.matrix(coef_by_species[, -1])  # drop species column
rownames(coef_mat) <- coef_by_species$Species
species_contrib <- which(coef_mat > 0.1, arr.ind = TRUE)
species_contrib <- aggregate(
    row ~ col,
    data = species_contrib,
    FUN = function(ix) {
        species <- rownames(coef_mat)[ix]
        species <- substr(species, 1, 3)  # abbreviate species names
        paste(species, collapse = ".")
    }
)

anames(fit_iris) <- species_contrib$row
anames(fit_iris)
```

```{r iris-palette, include = FALSE}
pal_aa <- c(set = "#E69F00", set.ver = "#E7E300", vir = "#009E73")
pal_aa <- pal_aa[anames(fit_iris)]  # reorder to match archetype order
```

### Archetype feature profiles

`what = "profiles"` shows a grouped barplot of the archetype coordinates on
the original feature scale. All other arguments are passed to `barplot()`.

```{r iris-profiles, fig.cap = "Archetype feature profiles. Each group of bars is one feature; colours distinguish archetypes."}
plot(fit_iris, what = "profiles",
     horiz = TRUE,
     names.arg = sub("\\.", "\n", colnames(iris[, 1:4])),
     las = 1,
     col = pal_aa,
     xlab = "Feature value (original scale)", ylab = "",
     main = "Archetype feature profiles")
```

### Sample compositions

With named archetypes, we can explore how all samples are distributed in
composition space.

```{r iris-simplex, fig.cap = "Ternary plot of iris compositions, coloured by species. Each point is a sample placed by its three archetype weights."}
plot(fit_iris, what = "simplex",
     col = pal_sp[as.character(iris$Species)], pch = 19,
     pca = TRUE, col.pca = "black", robust = FALSE)
legend("topright", legend = names(pal_sp), col = pal_sp, pch = 19, bty = "n")
```

The ternary structure shows a clear separation of species in composition space:
setosa concentrates near one vertex, virginica near another, and versicolor
occupies intermediate mixtures along the lower edge (small setosa contribution).
The black curved line is the PCA trajectory projected into the simplex,
summarizing the dominant continuous transition from setosa-like to
virginica-like compositions through the hybrid archetype of setosa+versicolor.

We can also see from the ternary plot that even though on average more
versicolor samples are near the hybrid archetype, the setosa contribution is
stronger because AA looks for "extreme" observations near the boundary of the
data hull and there lies one setosa "outlier". This illustrates a general property
of AA: archetypes anchor at the geometric boundary, so even a single extreme
observation can pull a vertex toward it. For analyses where outliers are a
concern, see <a href="#robust-fitting">*Robust fitting*</a>.

The composition barplot also nicely shows the transition from virginica-like to
setosa-like samples, with the hybrid archetype sitting in between.

```{r iris-compositions, fig.cap = "Composition barplot: each bar is one iris sample. Single-colour bars are near-pure specimens; blended bars sit between archetypes."}
plot(fit_iris, what = "compositions",
     main = "Archetype compositions across iris samples",
     cluster_rows = "PC1", cluster_cols = FALSE, col = pal_aa
)
```

### Predicting compositions for new observations

Once a model is fitted, `predict()` maps new samples into the archetype
simplex without re-fitting.

```{r predict}
new_obs <- matrix(
  c(5.1, 3.5, 1.4, 0.2,   # setosa-like
    6.7, 3.0, 5.2, 2.3),  # virginica-like
  nrow = 2, byrow = TRUE,
  dimnames = list(c("obs1", "obs2"), colnames(iris[, 1:4]))
)
print(new_obs)

predict(fit_iris, newdata = new_obs)
```

By default, each predicted row is non-negative and sums to 1 (`type = "compositions"`).
If you want "denoised" values in the original feature space, request a reconstruction:

```{r predict-reconstruction}
predict(fit_iris, newdata = new_obs, type = "reconstruction")
```

## Practical considerations

### Scaling

| Situation                                 | `scale` value            |
|-------------------------------------------|--------------------------|
| Mixed-unit features (default)             | `TRUE` — column z-score  |
| Features already on their "natural" scale | `FALSE`                  |
| Custom feature importance                 | Named numeric vector     |

The decision of whether to scale the data is similar to that in PCA: if the features
are on different scales, it is usually advisable to scale them to give them
equal weight in the analysis. Otherwise, the optimization process may effectively
ignore them. Custom scaling by passing a vector of scale factors is also
supported similar to how `scale()` works in base R (by dividing the columns).

Besides `scale`, `run_aa` also has an `sd_threshold = 1e-6` argument to remove
features with near-zero variance, which can cause numerical instability and
convergence issues. During reconstruction, the removed features are filled with
the mean value of the original data.

Finally, when some samples are more reliable/informative than others, weights
can be passed to `run_aa()` via the `weights` argument to give them more importance
in the calculation of the loss function.

### Multiple restarts

Because the AA objective is non-convex, the solution is sensitive to
the initial archetypes. To mitigate this, `run_aa()` supports multiple
independent restarts via the `nrep` argument. Setting `nrep > 1` only
returns the best solution across all runs, so the output is always a single
`archetypes` object.

The default initialization method is `"furthest_sum"`, which selects points
from the dataset such that the sum of their pairwise distances is maximized.
Because this places the initial archetypes near the boundary of the data, the
optimization starts from a reasonable configuration. The following small
experiment runs AA five times on the toy dataset from independent starting
configurations and compares the loss at the first and last iteration:

```{r furthest-sum-experiment}
runs <- do.call(rbind, lapply(1:5, function(seed) {
  set.seed(seed); fit_fs  <- run_aa(toy, K = 3L, nrep = 1L)
  set.seed(seed); fit_uni <- run_aa(toy, K = 3L, nrep = 1L, init = "uniform_archetypes")
  data.frame(
    run            = seed,
    fs_loss_start  = fit_fs$loss$loss[1],
    fs_loss_end    = tail(fit_fs$loss$loss, 1),
    uni_loss_start = fit_uni$loss$loss[1],
    uni_loss_end   = tail(fit_uni$loss$loss, 1)
  )
}))
knitr::kable(runs, digits = 3,
  caption = "Loss at first and last iteration for `furthest_sum` (fs) vs. `uniform_archetypes` (uni) across 5 seeds.")
```

`furthest_sum` consistently starts from a lower initial loss — because the
initializer already places candidate archetypes near the data boundary — and
the final losses are tightly clustered. `uniform_archetypes`, which selects
K observations uniformly at random, starts from higher and more variable initial
losses and may require a larger `nrep` to find the global minimum reliably.
See the <a href="initialization.html">**initialization vignette**</a> for a full comparison of all available
initializers.

### Convergence

Check `fit$converged` and inspect the loss trajectory:

```{r check-convergence, fig.cap = "RSS across iterations for the iris model."}
fit_iris$converged
tail(fit_iris$loss, 1)
plot(fit_iris, what = "loss", frame.plot = FALSE, type = "b",
     main = "Convergence of the iris model")
```

If the model has not converged, increase `max_iter` (default 100) or add more
restarts. If the loss decreases very slowly, then increasing the `max_iter` is
unlikely to help; consider increasing the `tol` or `tol_r2` parameter instead.
Note that this will not improve the fit, but it will save you some time and/or
avoid annoying warning messages by stopping the optimization earlier.

---

## Variants of Canonical AA and their implementation in YAAAP

### Relaxing the convex hull constraint with `delta`

By default (`delta = 0`) archetypes are constrained to the *convex hull* of
the data, i.e., they are strictly weighted averages of actual observations.

When the data is a **truncated** version of a larger pattern, the genuine
extremes are absent and the constrained model cannot recover the true vertices
and instead anchors to the most extreme observed points.

Setting `delta > 0` relaxes this constraint and allows archetypes to step
*outside* (or *inside*) the data hull [@Morup2012].  The following example makes this
concrete with synthetic data drawn from a **truncated equilateral simplex**:
any composition weight above 0.8 is removed, so no sample is close to any
true vertex.

```{r delta-data-fit, fig.cap = "Truncated simplex: Larger delta lets archetypes escape the data hull and approach the true positions."}
set.seed(20250804)
K_d <- 3L  # number of archetypes (simplex vertices)
M_d <- 2L  # number of features (2D for easy visualization)
N_init <- 1000L  # initial number of samples before truncation
thresh <- 0.8    # truncation threshold on composition weights

# True archetypes: equilateral triangle
A_true <- matrix(
    c(cos(seq(0, 2*pi*(K_d-1)/K_d, length.out = K_d)),
      sin(seq(0, 2*pi*(K_d-1)/K_d, length.out = K_d))),
    nrow = K_d, ncol = M_d
)

# Dirichlet-like mixing weights; remove observations near any vertex
S_raw   <- -log(matrix(runif(K_d * N_init), nrow = N_init, ncol = K_d))
S_raw   <- S_raw / rowSums(S_raw)
S_trunc <- S_raw[rowSums(S_raw > thresh) == 0, ]
X_delta <- S_trunc %*% A_true
print(nrow(X_delta))  # observations after truncation

# Fit models with different delta values
delta_vals <- c(0, 0.25, 0.5)
res_delta  <- lapply(delta_vals, function(d) archetypes_pgd(X_delta, K = K_d, delta = d))

# Plot all models together with the true archetypes
ix   <- c(1:3, 1)   # close the triangle
cols <- c("steelblue", "darkorange", "firebrick")

plot(A_true[ix, 1], A_true[ix, 2],
     type = "l", col = "black", lwd = 2, lty = 3,
     frame.plot = FALSE, xlab = "", ylab = "",
     xaxt = "n", yaxt = "n", asp = 1,
     main = "Effect of delta on recovered archetypes")
points(X_delta, pch = 19, cex = 0.4, col = "grey50")
for (i in seq_along(res_delta)) {
  lines(res_delta[[i]]$coordinates[ix, 1],
        res_delta[[i]]$coordinates[ix, 2],
        col = cols[i], lwd = 2)
}
legend("topright",
       legend = c("True simplex", paste0("\u03b4 = ", delta_vals)),
       col = c("black", cols), lty = c(3, 1, 1, 1), lwd = 2, bty = "n")
```

With `delta = 0` the estimated triangle is visibly smaller than the true one and
its vertices all lie within the data cloud.  Increasing `delta` allows
archetypes to reach outside the cloud towards the true vertices.  Use `delta`
with care: very large values allow archetypes to drift far from the data, losing
their interpretability as extreme prototypes.

### Robust fitting

Standard AA minimises squared residuals, making it sensitive to outliers.
Setting `robust = TRUE` applies **Tukey bisquare** row reweighting^[similar to M-estimators in `MASS::rlm`]:
observations with large residuals receive reduced influence at each iteration,
so outliers cannot monopolise an archetype.

Below we augment the `iris` dataset with five outliers and compare the results of
standard and robust AA.

```{r robust-data}
# Append five outlier observations to iris

add_outliers <- function(X, N, displacement) {
  x_mean <- colMeans(X) + displacement # place outliers at a distance from the mean
  x_sd <- apply(X, 2, sd)
  X_out <- rnorm(N * ncol(X), x_mean, x_sd)  # recycle x_mean, x_sd
  rbind(X, matrix(pmax(X_out, .1), nrow = N, byrow = TRUE))
}

set.seed(42L)

N_out <- 5L
displacement <- c(5, 0, 5, 0)  # displace in the first and third features
X <- as.matrix(iris[, 1:4]) |> add_outliers(N = N_out, displacement = displacement)
fit_std <- run_aa(X, K = 3L, robust = FALSE, tol = 1e-4)
fit_rob <- run_aa(X, K = 3L, robust = TRUE, tol = 1e-4)
```

We observe first that in the standard AA, one archetype is pulled towards the
outliers, while in the robust AA the outliers are effectively ignored and the
archetypes remain near the original data structure.

```{r robust-compare, fig.cap = "PCA projection of iris with five added outliers (crosses). Standard AA (red) is distorted towards the outlier cluster; robust AA (blue) closely recovers the reference triangle fitted on the clean data (dashed black)."}
pca <- prcomp(X, scale. = TRUE)
A_std <- predict(pca, fit_std[["coordinates"]])
A_rob <- predict(pca, fit_rob[["coordinates"]])
A_ori <- predict(pca, fit_iris[["coordinates"]])   # reference: fitted on clean iris
ix <- c(1:3, 1)
all_xy <- rbind(pca$x, A_std, A_rob, A_ori)
xl <- range(all_xy[, 1]); xl <- xl + diff(xl) * c(-0.07, 0.07)
yl <- range(all_xy[, 2]); yl <- yl + diff(yl) * c(-0.07, 0.07)
plot(pca$x[seq_len(nrow(iris)), ], col = "grey60", pch = 19, cex = 0.5,
     main = "PCA projection of iris data with outliers",
     xlim = xl, ylim = yl, frame.plot = FALSE, asp = 1)
points(pca$x[-seq_len(nrow(iris)), ], col = "black", pch = 4, cex = 1.5)
lines(A_ori[ix, ], col = "black", lwd = 1.5, lty = 2)
points(A_std, col = "firebrick", pch = 17, cex = 1.5)
lines(A_std[ix, ], col = "firebrick", lwd = 2)
points(A_rob, col = "steelblue", pch = 17, cex = 1.5)
lines(A_rob[ix, ], col = "steelblue", lwd = 2)
legend("topright",
       legend = c("Data", "Outliers", "Reference AA", "Standard AA", "Robust AA"),
       col    = c("grey60", "black", "black", "firebrick", "steelblue"),
       pch    = c(19, 4, NA, 17, 17),
       lty    = c(NA, NA, 2, 1, 1),
       bty    = "n")
```

Comparing the robust triangle to the reference (dashed), the archetypes closely
recover the positions found on the clean data, confirming that robust AA limits
the outliers' influence effectively.

```{r robust-compare-compositions, fig.width = 9, fig.height=5, fig.cap= "Ternary plot of iris compositions with and without robust fitting. Standard AA results are skewed towards the outliers, while robust AA maintains a more balanced representation of the original data structure."}
op <- par(mfrow = c(1, 2))
pal_sp["outlier"] <- "gray30"
x_species <- c(as.character(iris$Species), rep("outlier", N_out))
plot(fit_std, what = "simplex", col = pal_sp[x_species], pch = 19)
title(main = "Standard AA")
plot(fit_rob, what = "simplex", col = pal_sp[x_species], pch = 19)
title(main = "Robust AA")
par(fig = c(0, 1, 0, 1), new = TRUE)
legend("center", legend = names(pal_sp), col = pal_sp, pch = 19,
       bty = "n", xpd = NA)
```

In the robust simplex (right panel) the qualitative structure of the original data is
preserved: setosa occupies one pure vertex, virginica sits at the opposite corner, and
versicolor bridges the gap through the near-setosa extreme — the same three-way partition
seen in the clean-data iris model. In the standard simplex (left panel) one archetype
has been pulled toward the outlier cluster; the remaining iris samples are spread across
a distorted triangle and the clean species separation is lost.

When `robust = TRUE`, the tuning constant `tukey_c` (default 4.685) controls the breakdown point;
smaller values downweight moderately large residuals more aggressively.

### Missing data

When `X` contains `NA` values, `run_aa()` automatically switches to the
missing-data PGD objective (`missing = TRUE`): only the *observed* entries
contribute to the gradient, and the loss is computed over those entries only.

```{r missing-data}
# Introduce ~10 % missingness at random
set.seed(42L)

X <- as.matrix(iris[, 1:4])
X_miss <- X
X_miss[sample(length(X_miss), size = round(0.1 * length(X_miss)))] <- NA

set.seed(1L)
fit_miss <- run_aa(X_miss, K = 3L, nrep = 3L)   # missing = TRUE auto-detected
fit_miss
```

We see that there is a warning of non-convergence. However, the final $R^2$ is
good `r round(tail(fit_miss$loss$r2, 1), 3)`, suggesting that the fit is still
reasonable.

We can reconstruct the missing entries from the fitted model and check how well
they match the original data.

```{r missing-fitted}
X_hat_miss <- fitted(fit_miss)
any(is.na(X_hat_miss))   # FALSE = all positions reconstructed

# compare reconstructed vs original values at missing positions
idx_miss <- is.na(X_miss)
cor(X_hat_miss[idx_miss], X[idx_miss])
```

Next, we examine how the missing values affect the archetype positions in feature space.

```{r missing-compare, fig.width = 6, fig.height=6, fig.cap = "Archetype positions with (red) and without (blue) missing data"}
A <- rbind(fit_miss[["coordinates"]], fit_iris[["coordinates"]])
pairs(rbind(X, A),
      upper.panel = NULL, asp = 1,
      col = c(rep("grey60", nrow(X)), rep(c("firebrick", "steelblue"), each = 3)),
      pch = rep(c(19, 17), c(nrow(X), nrow(A))), cex = rep(c(.5, 1.5), c(nrow(X), nrow(A))),
      main = "Archetype coordinates")
legend("topright", inset = c(0.02,.15),
       legend = c("Data", "AA with missing data", "AA complete data"),
       col = c("grey60", "firebrick", "steelblue"), pch = c(19, 17, 17), bty = "n")
```

The archetypes from the missing-data fit (red) are close to those from the complete-data fit (blue)
and they do not affect the geometry of the simplex much.

Note that for a single archetype object `plot(fit_miss, what = "coordinates")`
will produce the same plot as above but without the complete-data archetypes for
comparison that is why we reconstructed the plot manually here.

This vignette covers only the Euclidean setting. YAAAP also supports AA on
**directional** and **compositional** data; these variants are demonstrated in
the *Non-Euclidean Geometries* vignette.

---

## Algorithm details

### Algorithmic approaches: NNLS vs PGD

The CRAN package
[`archetypes`](https://cran.r-project.org/package=archetypes) implements AA
via **alternating NNLS** (non-negative least squares): at each step a
constrained NNLS subproblem is solved via `nnls::nnls()`.  YAAAP's
`method = "nnls"` reproduces this approach and is available for compatibility.

The default `method = "pgd"` uses **projected gradient descent** instead.
PGD skips the inner NNLS solves working directly on the gradient of the
reconstruction error.

Unlike NNLS, which solves a constrained subproblem at each step, PGD works
entirely with matrix multiplications, avoiding the overhead of an inner solver.
The following benchmark on the toy dataset with 10 restarts illustrates this
difference:

```{r benchmark}
toy_mat    <- as.matrix(toy)
nrep_bench <- 10L
seeds      <- seq_len(nrep_bench)

run_single <- function(method, seed) {
  set.seed(seed)
  fit <- run_aa(toy_mat, K = 3L, nrep = 1L, method = method)
  c(n_iter = nrow(fit$loss), final_loss = norm(resid(fit), "F")^2)
}

t_pgd  <- system.time(res_pgd  <- vapply(seeds, run_single, numeric(2), method = "pgd") )[["elapsed"]]
t_nnls <- system.time(res_nnls <- vapply(seeds, run_single, numeric(2), method = "nnls"))[["elapsed"]]

n_pgd  <- sum(res_pgd["n_iter", ])
n_nnls <- sum(res_nnls["n_iter", ])

win_pgd <- sum(res_pgd["final_loss", ] < res_nnls["final_loss", ])

bench <- data.frame(
  Method         = c("PGD", "NNLS"),
  `Total (s)`    = c(t_pgd, t_nnls),
  `Total iters`  = c(n_pgd, n_nnls),
  `ms / iter`    = 1000 * c(t_pgd / n_pgd, t_nnls / n_nnls),
  `Mean RSS`     = c(median(res_pgd["final_loss", ]), median(res_nnls["final_loss", ])),
  `Wins`         = c(win_pgd, nrep_bench - win_pgd),
  check.names    = FALSE
)
knitr::kable(bench,
             digits = 2,
             caption = sprintf(
               "Benchmark: %d single-restart runs on %d \u00d7 %d toy dataset.",
               nrep_bench, nrow(toy_mat), ncol(toy_mat)))
```

Also, because PGD works directly on the gradient, it can be easily adapted
to handle extra terms like the slack terms used in the `delta` relaxation
and missing residuals in the `missing = TRUE` objective. Both methods
can handle the `robust = TRUE` fitting with Tukey bisquare reweighting,
as well as any custom weighting scheme passed via the `scale` and `weights`
arguments.

### PGD Algorithm

The algorithm is described in detail [@Morup2012] and is implemented in the
`archetypes_pgd()` function.  The main steps are:

1. **Initialisation**: selects K well-separated data
   points, giving good initial coverage of the data hull.
2. **S-step** — for fixed archetypes $A$, update compositions $S$ by a
   projected gradient step onto the probability simplex.
3. **B-step** — for fixed $S$, update the archetype generators $B$
   by a projected gradient step onto the row-stochastic simplex.
4. **a-step** (optional): in case of `delta > 0`, update the slacks `a`
   that allow archetypes $A$ to step outside the data hull defined by $B$.
5. **Repeat** until the relative change in RSS drops below `tol`, or $R^2$
   exceeds `tol_r2`.

When `pseudo_pgd = TRUE` (default), the gradient is modified before each step
so that it only redistributes weight among archetypes rather than changing their
total, keeping the proposed update close to the simplex surface. The step size
starts at `step_size` and is multiplied by `step_shrinkage` after each rejected
proposal; if no improvement is found after `max_iter_optimizer` reductions, the
step is skipped entirely.

If no update is accepted after `max_no_update` back-tracking steps, the
algorithm halts with a warning of non-convergence.

#### Key arguments

| Argument              | Default | Description                                           |
|-----------------------|---------|-------------------------------------------------------|
| `max_iter`            | 100     | Maximum outer iterations                              |
| `tol`                 | 1e-6    | Relative RSS change stopping criterion                |
| `tol_r2`              | 0.9999  | R² early-stop threshold                               |
| `nrep`                | 1       | Independent random restarts (best returned)           |
| `scale`               | `TRUE`  | Column z-scoring before fitting                       |
| `delta`               | 0       | Convex-hull relaxation (see <a href="#relaxing-the-convex-hull-constraint-with-delta">§ *delta*</a>)  |
| `pseudo_pgd`          | `TRUE`  | Remove radial component of gradient before step       |
| `step_size`           | 1.0     | Initial step size for the line search                 |
| `max_iter_optimizer`  | 10      | Maximum back-tracking steps per outer iteration       |
| `step_shrinkage`      | 0.5     | Shrinkage factor per back-track                       |
| `max_no_update`       | 5       | Consecutive no-improvement steps before early halt    |

#### Loss tracking

As introduced in the first example, the `$loss` data frame tracks one row per
logged iteration; the column meanings are summarised there. The full sequence
for the iris model is:

```{r loss-df}
head(fit_iris$loss, 6)
```

The loss metrics may differ from the direct computation (for example
`norm(X - fitted(fit_iris), "F")^2`) because the loss is computed on the
scaled data or other internal transformations. Thus, the `loss` metrics should
only be used for monitoring convergence and comparing different runs of the same
model type, not for comparing across different models or datasets. That is why
we use the generic term "loss" instead of "RSS" in the column name.

The condition numbers `k_S` and `k_A` are diagnostics of numerical stability and
are especially relevant for the case of `method = "nnls"` because it involves
solving linear systems with these matrices. High `k_S` or `k_A` values signal
numerical issues and are worth investigating with additional restarts or a
larger `delta`.  Most users will not need to worry about these.

---

## Conclusion

### Limitations and challenges

Some important practical considerations and limitations of AA:

- **Presence of (Poor) Local Minima**: The AA objective is non-convex,
  so the algorithm is not guaranteed to find the global minimum.
- **Initialization sensitivity**: Because of the presence of local minima,
  the choice of initial archetypes can influence the final solution.
- **Number of archetypes**: is the most important tuning parameter and its
  choice is subjective and can affect the interpretability and quality of the fit.
- **Sensitivity to Convex Hull**: Standard AA can be sensitive to outliers or
  misspecification when the true extremes lie outside the observed convex hull.
- **Euclidean assumption**: The standard AA formulation assumes a Euclidean
  geometry (Frobenius norm), which may not be appropriate for all data types.

Several methods have been proposed to mitigate these issues and this package
implements some of them. The most common strategies are:

- **Multiple restarts and initialisation**: To combat non-convexity and
  initialization sensitivity, it is common to run AA with multiple independent
  restarts (`nrep`) and select the best solution. Sophisticated initialization
  strategies (see <a href="initialization.html">**initialization vignette**</a>) can also avoid poor local minima
  and improve both solution quality and convergence speed.

- **Choosing the number of archetypes**: The number of archetypes `K` is the
  most important tuning parameter. The choice can be guided by domain knowledge,
  cross-validation, or information criteria, but it remains a subjective decision
  that affects the interpretability and quality of the fit. This package provides
  tools to help with this choice; see **tuning vignette**.

- **Robust fitting**: Because archetypes are constrained to be convex combinations
  of the data, an outlier can attract an entire archetype to itself, leading to a
  potentially misleading result. The option `robust = TRUE` applies Tukey bisquare
  reweighting (tuned by `tukey_c`) to reduce the influence of such observations
  (see <a href="#robust-fitting">*Robust fitting*</a>).

- **Delta relaxation**: If the genuine extremes of the underlying data-generating
  process are not observed, standard AA will be unable to recover them. To
  accommodate this situation, the <a href="#relaxing-the-convex-hull-constraint-with-delta">`delta` relaxation</a>
  allows archetypes to step outside the data hull and better capture true extremes
  when the data is truncated.

- **Non-Euclidean geometries**: The standard AA formulation assumes a Euclidean
  geometry (Frobenius norm), which may not be appropriate for all data types.
  Extensions of AA to other geometries and data types are presented in the
  **"Non-Euclidean Geometries" vignette**.

### Summary

| Task                          | Function / accessor            |
|-------------------------------|--------------------------------|
| Fit a model (matrix)          | `run_aa(X, K, ...)`            |
| Fit a model (formula)         | `run_aa(y ~ ., data, K, ...)`  |
| Extract compositions          | `fit$compositions`             |
| Extract coordinates           | `fit$coordinates`              |
| Extract coefficients          | `fit$coefficients`             |
| Reconstruct data              | `fitted(fit)`                  |
| Residuals                     | `residuals(fit)`               |
| Predict new samples           | `predict(fit, newdata)`        |
| Rename archetypes             | `anames(fit) <- c(...)`        |
| Loss / convergence            | `fit$loss`, `fit$converged`    |
| Feature profile plot          | `plot(fit, "profiles")`        |
| Composition barplot           | `plot(fit, "compositions")`    |
| Simplex plot                  | `plot(fit, "simplex")`         |
| Coordinate scatter            | `plot(fit, "coordinates")`     |
| Convergence plot              | `plot(fit, "loss")`            |
| Robust fitting                | `run_aa(..., robust = TRUE)`   |
| Missing-data fitting          | `run_aa(...)` (auto-detects)   |


## Session info

```{r session-info}
sessionInfo()
```

## References
