---
title: "Non-Gaussian and Metric Variants of Archetypal Analysis"
bibliography: ../references.bib
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Non-Gaussian and Metric Variants of Archetypal Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5.5,
    fig.align = "center",
    warning = FALSE
)
set.seed(42)
par(bty = "n")
library(yaap)
```

## One simplex, several geometries

Classical archetypal analysis (AA) represents each observation as a convex
mixture of a small number of extreme profiles [@Cutler1994]. In yaap, the main
objects keep the same interpretation across most variants:

- `coefficients` describe each archetype as a mixture of training observations
  or training observation profiles.
- `compositions` describe each observation as a mixture of archetypes.
- `coordinates` are the fitted archetype profiles, when an input-space profile
  is available.

What changes is the notion of closeness. Ordinary AA uses Euclidean squared
error. Metric and kernel AA keep a Gaussian squared-error view but alter the
inner-product geometry. Probabilistic AA replaces squared error with a
family-specific likelihood. Directional AA replaces Euclidean residuals with an
angular Watson-style loss for data where direction matters more than magnitude
[@Seth2016; @Olsen2022].

| Variant              | Main function                              | What changes                         | Typical data                                |
| -------------------- | ------------------------------------------ | ------------------------------------ | ------------------------------------------- |
| Metric Gaussian AA   | `run_aa(..., scale = G)`                   | Feature inner product                | Curves, spectra, correlated features        |
| Kernel AA            | `run_aa(..., method = "kernel")`           | Implicit feature-space inner product | Nonlinear geometry                          |
| Probabilistic AA     | `run_aa(..., method = "paa", family = ...)` | Likelihood/objective                 | Binary, count, compositional observations   |
| Directional AA       | `run_aa(..., method = "directional")`      | Angular objective                    | Unit directions, polarity-invariant signals |

This vignette walks through those variants in that order. The examples are
small and synthetic so they can be re-run quickly.

```{r helpers, include = FALSE}
pal_aa <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7")
pal_data <- "grey70"

sample_simplex <- function(n, k) {
    Z <- matrix(stats::rexp(n * k), nrow = n, ncol = k)
    Z / rowSums(Z)
}

final_loss <- function(fit) {
    tail(fit[["loss"]], 1)
}

unit_rows <- function(X) {
    X <- as.matrix(X)
    X / sqrt(rowSums(X * X))
}

plot_proxy_archetypes <- function(X, A_linear, A_kernel, main) {
    plot(
        X,
        asp = 1, pch = 19, cex = 0.45, col = pal_data,
        xlab = "x", ylab = "y", main = main
    )
    points(A_linear, pch = 4, cex = 1.25, lwd = 1.5, col = "black")
    points(A_kernel, pch = 1, cex = 1.45, lwd = 1.6, col = pal_aa[2])
}
```

## Metric Gaussian AA

Some data are non-Gaussian in representation but can still be reduced to
ordinary Gaussian AA after choosing the right geometry. Functional data are a
common example: if rows are smooth curves sampled on a grid, then a pointwise
Euclidean norm treats all grid locations as equally spaced and independent. A
more appropriate approximation to an integrated squared error uses a metric on
the feature axis.

In yaap the two sides of this weighted Gaussian objective are controlled by
separate arguments. Let $E = X - SA$ be the residual matrix, let
$W = \mathrm{diag}(w_1, \ldots, w_N)$ be a diagonal matrix of sample weights, and
let $G$ be the feature metric. The fitted objective is conceptually

$$
  \mathcal{L}(S,A; W, G)
    = \mathrm{tr}\{W^{1/2} E G E^\top W^{1/2}\}
    = \mathrm{tr}\{W E G E^\top\}.
$$

The `weights` argument sets $W$; when it is omitted, $W$ is the identity matrix.
The `scale` argument sets $G$. Conceptually, every non-kernel Gaussian use of
`scale` is a choice of this feature metric: `scale = FALSE` uses the identity
metric, a numeric vector gives a diagonal metric through per-feature scale
factors, and a positive-definite matrix supplies a full non-diagonal metric. Use
`weights`, not `scale`, when the goal is to make some samples count more than
others.

```{r metric-data}
t_grid <- seq(0, 1, length.out = 60)
A_true <- rbind(
    early_peak = exp(-45 * (t_grid - 0.18)^2),
    late_peak  = exp(-45 * (t_grid - 0.82)^2),
    ramp       = 0.2 + 0.8 * t_grid
)

S_true <- sample_simplex(90, 3)
X_fun <- S_true %*% A_true +
    matrix(rnorm(90 * length(t_grid), sd = 0.025), nrow = 90)
colnames(X_fun) <- sprintf("t%02d", seq_along(t_grid))
```

```{r metric-fit}
feature_weight <- ifelse(t_grid < 0.25 | t_grid > 0.75, 3, 0.5)
G_diag <- diag(feature_weight)  # diagonal feature metric G

set.seed(1)
fit_plain <- run_aa(
    X_fun,
    K = 3,
    scale = FALSE,  # G = I
    max_iter = 25,
    tol_r2 = 1
)

set.seed(1)
fit_metric <- run_aa(
    X_fun,
    K = 3,
    scale = G_diag,  # G = G_diag
    max_iter = 25,
    tol_r2 = 1
)

rbind(
    plain_metric = final_loss(fit_plain),
    weighted_metric = final_loss(fit_metric)
)
```

```{r metric-plot, echo=FALSE, fig.cap = "Synthetic curves and metric-weighted archetype profiles. The shaded bands mark regions emphasized by the diagonal feature metric."}
sample_ix <- seq(1, nrow(X_fun), length.out = 25)

matplot(
    t_grid,
    t(X_fun[sample_ix, ]),
    type = "l", lty = 1, col = adjustcolor("black", 0.12),
    xlab = "t", ylab = "Value", main = "Metric Gaussian AA"
)
usr <- par("usr")
rect(0, usr[3], 0.25, usr[4], col = adjustcolor(pal_aa[2], 0.12), border = NA)
rect(0.75, usr[3], 1, usr[4], col = adjustcolor(pal_aa[2], 0.12), border = NA)
matlines(
    t_grid,
    t(coordinates(fit_metric)),
    lty = 1, lwd = 2, col = pal_aa[1:3]
)
legend(
    "top",
    legend = anames(fit_metric),
    lty = 1, lwd = 2, col = pal_aa[1:3],
    bty = "n", horiz = TRUE
)
```

The diagonal metric above is just a simple quadrature-like weighting. A full
positive-definite matrix can encode correlated features, basis inner products,
or a functional-data penalty. The `run_aa.fd()` method uses this idea internally:
it fits AA to basis coefficients with the basis inner-product matrix. The caveat
is that this remains Gaussian AA. The data are still reconstructed by
`compositions %*% coordinates`; only the metric used to judge residuals has
changed.

## Kernel AA

Kernel AA is the nonlinear analogue of changing the inner product. Instead of
choosing an explicit feature metric, it works with a Gram matrix
$K_{ij} = k(x_i, x_j)$ and fits archetypes in the kernel-induced feature space
[@Morup2012]. With the linear kernel, this recovers the ordinary Euclidean
geometry. With nonlinear kernels, the convex hull is formed after the implicit
feature map.

The example below has a dense core and an outer ring. Linear AA sees the input
space and places proxy archetypes around broad outer directions. A Laplace
kernel changes the geometry so local neighborhoods are emphasized.

```{r kernel-data}
set.seed(2011)
n_outer <- 120
n_inner <- 80

theta_outer <- runif(n_outer, 0, 2 * pi)
r_outer <- 1 + 0.12 * runif(n_outer)
X_outer <- cbind(r_outer * cos(theta_outer), r_outer * sin(theta_outer))

theta_inner <- runif(n_inner, 0, 2 * pi)
r_inner <- 0.15 * runif(n_inner)
X_inner <- cbind(r_inner * cos(theta_inner), r_inner * sin(theta_inner))

X_ring <- rbind(X_outer, X_inner)
colnames(X_ring) <- c("x", "y")
```

```{r kernel-fit}
set.seed(2)
fit_linear <- run_aa(
    X_ring,
    K = 8,
    scale = FALSE,
    max_iter = 35,
    tol_r2 = 1
)

set.seed(2)
fit_kernel <- run_aa(
    X_ring,
    K = 8,
    method = "kernel",
    kernel = "laplace",
    kernel_args = list(sigma = 0.35),
    max_iter = 35,
    tol_r2 = 1
)

rbind(
    linear = final_loss(fit_linear),
    laplace_kernel = final_loss(fit_kernel)
)
```

```{r kernel-plot, echo=FALSE, fig.cap = "Linear AA archetypes and nonlinear kernel-AA input-space proxy archetypes on a ring/core dataset."}
plot_proxy_archetypes(
    X_ring,
    coordinates(fit_linear),
    coordinates(fit_kernel),
    main = "Linear AA vs kernel-AA proxies"
)
legend(
    "topright",
    legend = c("Data", "Linear AA", "Kernel-AA proxy"),
    pch = c(19, 4, 1),
    col = c(pal_data, "black", pal_aa[2]),
    pt.cex = c(0.7, 1.2, 1.4),
    bty = "n"
)
```

For nonlinear kernels, `coordinates` are input-space proxy coordinates computed
as `coefficients %*% data`. They are useful for plotting, but they are not exact
preimages of the Hilbert-space archetypes. This distinction also explains why
`fitted()` and out-of-sample `predict()` are currently not defined for
`kernel_archetypes`: new samples would require kernel-specific cross-Gram
evaluation and a reconstruction convention for the selected kernel.

The kernel choice and bandwidth are substantive modeling decisions. Very local
kernels can make many points effectively extreme in feature space, a phenomenon
also discussed in kernel frame methods [@Mair2018Frame]. In practice, compare a
small grid of kernel parameters and inspect whether the resulting compositions
are stable.

## Probabilistic AA

PAA changes the objective rather than the inner product. It keeps the simplex
structure, but archetypes are profiles in the parameter space of an observation
family [@Seth2016]. In yaap, the supported families are `"gaussian"`,
`"bernoulli"`, `"poisson"`, and `"multinomial"`.

The following binary example represents observations as answers to six yes/no
items. The archetypes are probability profiles, not binary observations.

```{r paa-bernoulli-data}
set.seed(3)
P_true <- rbind(
    broad_yes = c(0.85, 0.80, 0.75, 0.20, 0.15, 0.10),
    middle    = c(0.25, 0.75, 0.35, 0.70, 0.35, 0.65),
    broad_no  = c(0.10, 0.15, 0.20, 0.80, 0.85, 0.75)
)
S_bin <- sample_simplex(120, 3)
P_bin <- S_bin %*% P_true
X_bin <- matrix(
    rbinom(length(P_bin), size = 1, prob = as.vector(P_bin)),
    nrow = nrow(P_bin)
)
colnames(X_bin) <- paste0("item_", seq_len(ncol(X_bin)))
```

```{r paa-bernoulli-fit}
set.seed(4)
fit_bernoulli <- run_aa(
    X_bin,
    K = 3,
    method = "paa",
    family = "bernoulli",
    max_iter = 35,
    tol_r2 = 1
)

fit_bernoulli
```

```{r paa-profile, echo=FALSE, fig.cap = "Bernoulli PAA archetypes. Bar heights are fitted response probabilities, not observed 0/1 data."}
plot(
    fit_bernoulli,
    what = "profiles",
    col = pal_aa[1:3],
    ylim = c(0, 1),
    main = "Bernoulli archetype probabilities"
)
```

Prediction keeps the learned archetypes fixed and returns reconstructions by
default. With `type = "compositions"`, it returns the new composition weights.

```{r paa-predict}
S_new <- predict(fit_bernoulli, X_bin[1:5, ], type = "compositions", max_iter = 20)
round(S_new, 3)

P_new <- predict(
    fit_bernoulli,
    X_bin[1:5, ],
    max_iter = 20
)
round(P_new, 2)
```

For count data, use `family = "poisson"`; for rows of category counts, use
`family = "multinomial"`. Multinomial fitted values preserve each row total.

```{r paa-multinomial}
set.seed(5)
Topic_true <- rbind(
    visual = c(0.55, 0.25, 0.10, 0.05, 0.05),
    text   = c(0.05, 0.10, 0.55, 0.25, 0.05),
    mixed  = c(0.15, 0.25, 0.15, 0.20, 0.25)
)
S_txt <- sample_simplex(80, 3)
Theta <- S_txt %*% Topic_true
totals <- sample(40:80, nrow(Theta), replace = TRUE)
X_txt <- t(vapply(seq_len(nrow(Theta)), function(i) {
    as.integer(rmultinom(1, totals[i], Theta[i, ]))
}, integer(ncol(Theta))))
colnames(X_txt) <- paste0("term_", seq_len(ncol(X_txt)))

set.seed(6)
fit_multi <- run_aa(
    X_txt,
    K = 3,
    method = "paa",
    family = "multinomial",
    max_iter = 25,
    tol_r2 = 1
)

head(round(fitted(fit_multi), 1), 3)
rowSums(fitted(fit_multi)[1:3, ])
rowSums(X_txt[1:3, ])
```

PAA's main caveat is interpretive: the convex hull lives in parameter space.
For non-Gaussian families, `plot(fit, what = "coordinates")` is intentionally
unavailable because the observed data and fitted archetypes are not in the same
space. PAA also does not currently support sample `weights`, robust fitting, or
missing-data fitting.

## Directional AA

Directional AA changes the objective to model directions rather than Euclidean
magnitudes. It is useful when rows lie on, or are naturally normalized to, a
sphere and signs may be arbitrary. The implementation uses a Watson-style loss
that is insensitive to polarity after hemisphere handling [@Olsen2022].

The synthetic example below draws points from a spherical triangle, then flips
some rows by multiplying them by `-1`. Euclidean AA treats the flipped rows as
opposite points. Directional AA can model them as the same direction.

```{r directional-data}
set.seed(7)
A_dir_true <- rbind(
    c(1, 0, 0),
    c(0, 1, 0),
    c(0, 0.15, 1)
)
S_dir <- sample_simplex(120, 3)
X_dir <- unit_rows(S_dir %*% A_dir_true)
flip <- sample(c(-1, 1), nrow(X_dir), replace = TRUE)
X_dir_flip <- X_dir * flip
colnames(X_dir_flip) <- c("x", "y", "z")
```

```{r directional-fit}
set.seed(8)
fit_euclidean_dir <- run_aa(
    X_dir_flip,
    K = 3,
    scale = FALSE,
    max_iter = 30,
    tol_r2 = 1,
    max_kappa = Inf,
    sd_threshold = 0
)

set.seed(8)
fit_directional <- run_aa(
    X_dir_flip,
    K = 3,
    method = "directional",
    max_iter = 30,
    tol_r2 = 1,
    max_kappa = Inf
)

rbind(
    euclidean = final_loss(fit_euclidean_dir),
    directional = final_loss(fit_directional)
)
```

```{r directional-directions}
round(fit_directional[["directions"]], 3)
round(rowSums(fitted(fit_directional)^2), 6)[1:6]
```

```{r directional-plot, echo=FALSE, fig.cap = "First two coordinates of polarity-flipped spherical data. Directional AA estimates unit directions; Euclidean AA is pulled by sign-flipped observations."}
plot(
    X_dir_flip[, 1:2],
    asp = 1, pch = 19, cex = 0.5,
    col = ifelse(
        flip > 0,
        adjustcolor("black", 0.25),
        adjustcolor(pal_aa[2], 0.25)
    ),
    xlab = "x", ylab = "y", main = "Directional vs Euclidean AA"
)
points(
    unit_rows(coordinates(fit_euclidean_dir))[, 1:2],
    pch = 4, cex = 1.4, lwd = 1.5, col = "black"
)
points(
    fit_directional[["directions"]][, 1:2],
    pch = 16, cex = 1.2, col = pal_aa[1]
)
legend(
    "bottomleft",
    legend = c("Positive row", "Flipped row", "Euclidean AA", "Directional AA"),
    pch = c(19, 19, 4, 16),
    col = c(adjustcolor("black", 0.45), adjustcolor(pal_aa[2], 0.45),
            "black", pal_aa[1]),
    bty = "n"
)
```

Directional AA returns `directional_archetypes`, which extends the ordinary
`archetypes` class. `fitted()` returns row-normalized reconstructions aligned
to the original row polarity when the training data are available. The
directional loss is not comparable to Euclidean SSE, so compare directional
fits with directional diagnostics, not with ordinary residual sums of squares.

Current limitations are stricter than for Gaussian AA: input must be a dense
numeric matrix with no zero rows; missing data, robust fitting, and custom
`scale` values are not supported.

## Practical checklist

Use the simplest model that matches the data-generating assumptions:

- If a weighted or correlated feature metric is enough, stay with Gaussian AA
  and supply `scale`.
- If the structure is nonlinear but still geometric, try kernel AA and inspect
  sensitivity to kernel parameters.
- If observations are binary, counts, or category frequencies, use PAA and
  interpret archetypes as distribution parameters.
- If only direction matters, use directional AA and remember that magnitudes are
  not the object of fit.

Across all variants, use multiple seeds or `nrep` where supported. The
optimization problems remain non-convex, so stable conclusions should not rely
on a single initialization.

## References
