---
title: "Tidymodels Workflows with yaap"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Tidymodels Workflows with yaap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5,
    fig.align = "center",
    warning = FALSE
)
set.seed(42)
par(bty = "n")
```

## Overview

`yaap` fits archetypal analysis (AA) models. In a tidymodels workflow,
it is best understood as an unsupervised decomposition or feature-extraction
tool, closer in role to `recipes::step_pca()` than to a supervised `parsnip`
model specification. `yaap` learns archetype coordinates and sample
composition weights, then exposes those results through interfaces that work
naturally with the tidymodels ecosystem:

- `step_archetypes()` is a `recipes` step that learns archetypes during
  `prep()` and returns archetype composition weights during `bake()`.

This vignette uses only lightweight pieces of the ecosystem. Because AA is an
unsupervised representation method, the examples focus on preprocessing,
feature extraction, and model inspection rather than on `parsnip` fitting.

```{r libraries, message = FALSE}
library(yaap)
library(generics)
library(recipes)
library(tune)
```

The examples use the built-in `USArrests` data, the same data used in the
`recipes::step_pca()` examples. The four numeric variables are on different
scales, so the recipes below normalize them before fitting either PCA or AA.
State regions are used only for visualization.

```{r load-usarrests}
arrests <- USArrests
state <- rownames(arrests)
region <- state.region[match(state, state.name)]
region_cols <- c(
    "Northeast" = "#0072B2",
    "South" = "#D55E00",
    "North Central" = "#009E73",
    "West" = "#CC79A7"
)

head(data.frame(state = state, region = region, arrests, row.names = NULL))
table(region)
```

```{r plot-usarrests, fig.cap = "USArrests data: murder and assault rates by state region."}
plot(
    arrests$Murder, arrests$Assault,
    pch = 19, cex = 0.75,
    col = region_cols[as.character(region)],
    xlab = "Murder arrests per 100,000",
    ylab = "Assault arrests per 100,000",
    main = "USArrests",
    frame.plot = FALSE
)
legend(
    "topleft",
    legend = names(region_cols),
    col = region_cols,
    pch = 19,
    bty = "n"
)
```

## Choosing K with Scree, Elbow, and AIC

The main tuning parameter in archetypal analysis is `K`: the number of
archetypes. A traditional exploratory approach, also familiar from the classic
`archetypes` package, is to fit a sequence of models and inspect a screeplot or
elbow plot. The elbow is the smallest `K` after which reconstruction quality
improves only modestly.

`yaap` also provides an adapted AIC-like criterion for Euclidean Gaussian AA
models. This is a `yaap`-specific model-selection helper; it is not part of
the classic `archetypes` package workflow. It is useful for comparing candidate
`K` values on the same data, but it is still a model-selection aid rather
than a substitute for scientific judgment.

Here we fit `K = 1, ..., 6` to the normalized arrest variables. The
one-archetype baseline is useful for RSS and $R^2$, but the adapted AIC can be
undefined for `K = 1`, so we score the grid by RSS and compute AIC separately
where it is defined.

```{r k-selection-fit}
rss <- function(fit) {
    utils::tail(fit$loss$loss, 1L)
}

aic_or_na <- function(fit) {
    if (nrow(coordinates(fit)) == 1L) NA_real_ else AIC(fit)
}

arrests_scaled <- as.data.frame(scale(arrests))
rownames(arrests_scaled) <- state

set.seed(42)
k_values <- 1:6
fits_k <- lapply(k_values, function(k) {
    run_aa(
        arrests_scaled,
        K = k,
        init = "furthest_sum",
        max_iter = 100,
        tol_r2 = 0,
        max_kappa = Inf
    )
})

k_tbl <- data.frame(
    K = k_values,
    rss = vapply(fits_k, rss, numeric(1L)),
    r2 = vapply(fits_k, function(fit) utils::tail(fit$loss$r2, 1L), numeric(1L)),
    AIC = vapply(fits_k, aic_or_na, numeric(1L))
)
selected_ix <- which.min(k_tbl$AIC)
selected_K <- k_tbl$K[selected_ix]
fit <- fits_k[[selected_ix]]
analysis_K <- 3L

knitr::kable(
    k_tbl[, c("K", "rss", "r2", "AIC")],
    digits = c(0, 1, 3, 1)
)
```

```{r k-selection-plot, results = "hold", fig.width = 9, fig.height = 3.5, out.width = "100%", fig.cap = "RSS, unexplained variance, and adapted AIC across candidate values of K."}
op <- par(mfrow = c(1, 3), mar = c(4, 4, 2.5, 0.5), bty = "n")

plot(
    k_tbl$K, k_tbl$rss,
    type = "b", pch = 19,
    col = "steelblue4",
    frame.plot = FALSE,
    xlab = "Number of archetypes (K)",
    ylab = "Reconstruction loss",
    main = "RSS"
)
abline(v = selected_K, lty = 2, col = "firebrick")

plot(
    k_tbl$K, 1 - k_tbl$r2,
    type = "b", pch = 19,
    col = "steelblue4",
    frame.plot = FALSE,
    xlab = "Number of archetypes (K)",
    ylab = expression(1 - R^2),
    main = "Unexplained Variance"
)
abline(v = selected_K, lty = 2, col = "firebrick")

plot(
    k_tbl$K, k_tbl$AIC,
    type = "b", pch = 19,
    col = "steelblue4",
    frame.plot = FALSE,
    xlab = "Number of archetypes (K)",
    ylab = "Adapted AIC",
    main = "Information Criterion"
)
abline(v = selected_K, lty = 2, col = "firebrick")

par(op)
```

For these normalized variables, the adapted AIC favours `K = 2`. The recipe
examples below use `K = 3` as a visualization choice: it matches the familiar
three-component `step_pca()` example and lets each state be drawn on a regular
simplex by its three archetype composition weights.

```{r best-model}
k_tbl[selected_ix, c("K", "rss", "r2", "AIC")]
selected_K
analysis_K
```

In real analyses these plots should be treated as diagnostics, not as automatic
rules. AA optimization is non-convex, so single-restart curves can be noisy and
may not decrease monotonically for every larger `K`. Use more restarts and
domain knowledge when the elbow or information criterion is ambiguous.

## Using Archetypes in Recipes

`step_archetypes()` turns AA into a preprocessing step. During `prep()`, the
step fits archetypes on the selected numeric predictors. During `bake()`, it
projects data onto the learned archetype simplex and returns one composition
column per archetype.

The recipes argument is called `num_comp`, following tidymodels naming
conventions, but it is the same quantity as `K`.

```{r recipe-fit}
rec_base <- recipe(~ ., data = arrests) |>
    step_normalize(all_numeric())

rec <- rec_base |>
    step_archetypes(
        all_numeric(),
        num_comp = analysis_K,
        seed = 42,
        options = list(
            init = "furthest_sum",
            max_iter = 100,
            tol_r2 = 0,
            max_kappa = Inf
        )
    )

rec_prep <- prep(rec, training = arrests)
rec_prep
```

```{r recipe-bake}
baked <- bake(rec_prep, new_data = arrests)
head(data.frame(state = state, region = region, baked, row.names = NULL))

aa_cols <- paste0("A", seq_len(analysis_K))
range(rowSums(as.matrix(baked[, aa_cols])))
```

By default, the original numeric predictors are removed and replaced by
composition columns named `A1`, `A2`, and so on. Set
`keep_original_cols = TRUE` when both the normalized predictors and archetype
weights should be retained.

```{r recipe-keep-original}
rec_keep <- rec_base |>
    step_archetypes(
        all_numeric(),
        num_comp = analysis_K,
        keep_original_cols = TRUE,
        seed = 42,
        options = list(
            max_iter = 100,
            tol_r2 = 0,
            max_kappa = Inf
        )
    ) |>
    prep(training = arrests)

head(data.frame(
    state = state,
    region = region,
    bake(rec_keep, new_data = arrests),
    row.names = NULL
))
```

## Comparing `step_archetypes()` with `step_pca()`

The closest tidymodels analogy for `yaap` is `recipes::step_pca()`: both steps
learn an unsupervised representation of the numeric predictors during `prep()`.
The outputs have different meanings. PCA returns orthogonal score columns such
as `PC1`, `PC2`, and `PC3`; AA returns composition weights such as `A1`, `A2`,
and `A3` that are non-negative and sum to one for each state.

The two recipes below are trained in parallel on the same normalized arrest
variables.

```{r compare-pca-aa}
rec_aa <- rec_base |>
    step_archetypes(
        all_numeric(),
        num_comp = analysis_K,
        seed = 42,
        options = list(
            init = "furthest_sum",
            max_iter = 100,
            tol_r2 = 0,
            max_kappa = Inf
        )
    ) |>
    prep(training = arrests)

rec_pca <- rec_base |>
    step_pca(all_numeric(), num_comp = analysis_K) |>
    prep(training = arrests)

aa_scores <- bake(rec_aa, new_data = arrests)
pca_scores <- bake(rec_pca, new_data = arrests)

head(data.frame(
    state = state,
    region = region,
    pca_scores,
    aa_scores,
    row.names = NULL
))
```

The PCA step returns orthogonal score columns. The AA step returns convex
composition weights over three estimated archetypal states. Inspecting both
trained steps highlights the difference: PCA loadings describe directions of
variation, while AA coordinates describe extreme profiles in the normalized
feature space.

```{r compare-pca-aa-tidy}
tidy(rec_pca, number = 2, type = "variance")
tidy(rec_aa, number = 2)
```

```{r compare-pca-aa-plot, fig.width = 6, fig.height = 5, fig.cap = "PCA scores for USArrests, coloured by state region."}
rng <- extendrange(c(pca_scores$PC1, pca_scores$PC2))
plot(
    pca_scores$PC1, pca_scores$PC2,
    xlim = rng, ylim = rng,
    pch = 19, cex = 0.75,
    col = region_cols[as.character(region)],
    xlab = "PC1", ylab = "PC2",
    main = "step_pca()",
    frame.plot = FALSE
)
abline(h = 0, v = 0, col = "grey85")
legend(
    "topleft",
    legend = names(region_cols),
    col = region_cols,
    pch = 19,
    bty = "n"
)
```

The PCA plot is a rotated coordinate system for the original variables. The AA
representation keeps mixture weights over interpretable extremes. Those weights
can be inspected directly on the archetype simplex.

## Declaring Tunable Parameters

`step_archetypes()` declares `num_comp` and `delta` as tunable recipe
parameters. This lets the step participate in a larger tidymodels tuning
workflow when the rest of the modeling pipeline supplies resampling, metrics,
and an outcome.

```{r tunable}
rec_tune <- rec_base |>
    step_archetypes(
        all_numeric(),
        num_comp = tune(),
        delta = tune()
    )

tunable(rec_tune$steps[[2L]])
```

For supervised modeling pipelines, use `step_archetypes()` inside the broader
tidymodels workflow and tune `num_comp` alongside the downstream model
parameters.

## Regional Simplex View

For `K = 3`, the fitted AA compositions can be drawn on the regular simplex.
Each point is one state, placed by its three archetype weights and coloured by
Census region.

```{r simplex-region-plot, fig.width = 6, fig.height = 6, fig.cap = "USArrests archetype simplex, coloured by state region."}
aa_cols <- grep("^A[0-9]+$", names(aa_scores), value = TRUE)
aa_mat <- as.matrix(aa_scores[, aa_cols])
simplex_xy <- cbind(
    x = aa_mat[, "A2"] + 0.5 * aa_mat[, "A3"],
    y = sqrt(3) / 2 * aa_mat[, "A3"]
)

plot(
    simplex_xy,
    type = "n",
    xlim = c(0, 1),
    ylim = c(0, sqrt(3) / 2),
    asp = 1,
    axes = FALSE,
    xlab = "", ylab = "",
    main = "USArrests AA Simplex",
    frame.plot = FALSE
)
polygon(
    c(0, 1, 0.5),
    c(0, 0, sqrt(3) / 2),
    border = "grey55",
    col = NA,
    lwd = 1.2
)
text(
    c(-0.03, 1.03, 0.5),
    c(-0.03, -0.03, sqrt(3) / 2 + 0.04),
    labels = aa_cols,
    xpd = NA
)
points(
    simplex_xy,
    pch = 19,
    cex = 0.8,
    col = region_cols[as.character(region)]
)
legend(
    "topright",
    legend = names(region_cols),
    col = region_cols,
    pch = 19,
    bty = "n"
)
```

The simplex view answers a different question from PCA. PCA spreads states along
directions that explain variance in the standardized arrest rates. AA asks how
each state can be represented as a mixture of extreme arrest-rate profiles, so
near-corner states have more concentrated archetype weights and interior states
blend multiple profiles.
