---
title: "Tuning Archetypal Analysis with Information Criteria and Consistency"
bibliography: ../vignettes/references.bib
format:
  html:
    toc: true
    toc-depth: 3
    df-print: paged
    embed-resources: true
execute:
  collapse: true
  comment: "#>"
  fig-width: 7
  fig-height: 5
  fig-align: center
  warning: false
  message: false
params:
  mode: development
---

```{r setup, include = FALSE}
benchmark_budget_preset <- function(mode = c("development", "article", "vignette")) {
    mode <- match.arg(mode)
    presets <- list(
        development = list(
            mode = "development",
            n_scenarios = 2L,
            n_reps_data = 1L,
            n_restarts_per_K = 3L,
            max_iter = 60L,
            nrep_fit = 1L,
            K_extra = 2L,
            cache = TRUE
        ),
        article = list(
            mode = "article",
            n_scenarios = 4L,
            n_reps_data = 10L,
            n_restarts_per_K = 8L,
            max_iter = 150L,
            nrep_fit = 1L,
            K_extra = 3L,
            cache = TRUE
        ),
        vignette = list(
            mode = "vignette",
            n_scenarios = 1L,
            n_reps_data = 1L,
            n_restarts_per_K = 2L,
            max_iter = 40L,
            nrep_fit = 1L,
            K_extra = 1L,
            cache = FALSE
        )
    )
    presets[[mode]]
}

benchmark_budget <- benchmark_budget_preset(params$mode)

knitr::opts_chunk$set(
    cache = benchmark_budget$cache
)


load_yaap_for_article <- function() {
    candidates <- c(".", "..")
    descriptions <- file.path(candidates, "DESCRIPTION")
    is_yaap <- vapply(descriptions, function(path) {
        file.exists(path) && any(grepl("^Package: yaap$", readLines(path, n = 5L)))
    }, logical(1L))
    root <- candidates[which(is_yaap)[[1L]]]
    if (!is.na(root) && requireNamespace("pkgload", quietly = TRUE)) {
        pkgload::load_all(root, quiet = TRUE)
    } else {
        library(yaap)
    }
}


set.seed(42)
load_yaap_for_article()
```

## Goal

The main tuning parameter in archetypal analysis is `K`, the number of
archetypes. A common first diagnostic is a scree or elbow plot of reconstruction
loss across candidate values of `K`. `yaap` also provides information criteria
through `AIC()`, `BIC()`, and the lower-level `aa_ic()` helper.

This article expands that workflow into a small benchmark-style investigation.
The benchmark uses synthetic datasets where the true number of generating
archetypes is known. We compare two families of diagnostics:

- **Information criteria**: lower values indicate a better fit/parsimony tradeoff.
- **Consistency curves**: higher values indicate more stable recovered
  coordinates or sample compositions across repeated fits at the same `K`.

The article is not a CRAN vignette. It is intentionally placed under
`articles/` and can run heavier investigations locally.

## Computational Budget

All computational complexity is controlled by one object, `benchmark_budget`.
Render with a different preset by passing `params = list(mode = "...")`.

```{r budget}
data.frame(
    setting = names(benchmark_budget),
    value = unlist(benchmark_budget, use.names = FALSE),
    stringsAsFactors = FALSE
)
```

The supported presets are:

- `development`: fast settings for editing the article.
- `article`: deeper settings for a more informative non-CRAN report.
- `vignette`: small settings suitable if the source is later adapted into a
  CRAN-running vignette.

Every scenario, candidate `K` grid, repeated fit, and scoring step below reads
from `benchmark_budget`. This keeps the investigation easy to tune without
editing helper internals.

## Information Criteria

`aa_ic()` has two criterion types:

- `type = "aic"` uses a penalty multiplier of 2.
- `type = "bic"` uses a penalty multiplier of `log(n_eff)`.

The more important comparison here is the degrees-of-freedom estimate:

- `df_method = "suleman"` uses the adapted Suleman-style correction
  [@Suleman2017]. This is the default behind bare `AIC()` and `BIC()`.
- `df_method = "covariance"` treats the fitted weights as fixed and estimates
  the trace of the fitted-value smoother.
- `df_method = "parametric"` counts the full ambient simplex parameter count.
- `df_method = "active"` counts only active simplex-face dimensions, capped by
  local affine rank for archetype coordinates.

These criteria are not monotone in `K`; the selected model is the minimum of
the curve. Some criteria can be undefined for particular fits. For example, the
Suleman-style correction depends on covariance assumptions and may return `NA`
when those assumptions fail.

## Consistency Curves

Consistency asks a different question from AIC or BIC. Rather than balancing
fit and parsimony, it asks whether repeated fits at the same `K` recover similar
objects. `yaap` exposes this through `consistency()`.

For each candidate `K`, we fit multiple independent models and compute mean
pairwise consistency for:

- `what = "coordinates"`: are the archetype locations stable?
- `what = "compositions"`: are the sample barycentric weights stable?

Higher consistency is better, and the selected `K` is the maximum curve value.
Ties are resolved toward smaller `K`. This can favor stable underfit models, so
consistency should be read alongside IC and reconstruction diagnostics rather
than as a replacement for them.

For plotting, consistency values are shown on their raw scale after averaging
over data replicates. Coordinate and composition consistency therefore use their
native definitions rather than a shared max-normalized scale. Negative
coordinate-consistency values are left visible because they indicate matched
archetype locations farther apart than the data-scale variance reference.
locations farther apart than the data-scale variance reference.

## Benchmark Generation

Each synthetic dataset is generated from a known latent archetypal model. We
choose the true number of archetypes `K_true`, the feature dimension `M`, a
matrix of true archetype coordinates `A_true`, a Dirichlet concentration
parameter `alpha`, and a fixed Gaussian noise level `sigma`. For sample `n`,
the latent composition vector is sampled on the simplex and the observed row is
a noisy convex combination of the true archetypes:

$$s_n \sim \operatorname{Dirichlet}(\alpha_1, \ldots, \alpha_K),$$

$$e_n \sim \mathcal{N}(0, \sigma^2 I_M),$$

$$x_n = s_n A_\mathrm{true} + e_n.$$

Equivalently, for the full data matrix:

$$X = S_\mathrm{true} A_\mathrm{true} + E.$$

Rows of `S_true` are non-negative and sum to one, so the clean data are inside
the true archetype simplex. The noise matrix `E` perturbs those clean points in
feature space. The benchmark therefore separates several controls:

- `K_true` sets the ground-truth model size.
- `M` sets feature dimension.
- `A_true` sets archetype geometry and separation.
- `alpha` controls composition purity versus mixing.
- `corner_weights` can make one archetype sparsely represented.
- `sigma` controls observation noise independently of mixing.

```{mermaid}
flowchart LR
    alpha["α"] --> s["sₙ ~ Dirichlet(α)"]
    sigma["σ"] --> e["eₙ ~ N(0, σ² Iₘ)"]
    A["A_true"]

    subgraph plate["n = 1,...,N"]
        s
        e
        x["xₙ = sₙ A_true + eₙ"]
        s --> x
        e --> x
    end

    A --> x
    style plate fill:transparent,stroke:#6b7280,stroke-width:1.5px
```

The current "Choosing K" benchmark keeps `alpha` fixed within each scenario
because the focus is model-selection behavior over candidate `K`. A natural
future benchmark axis is an **Archetype Discoverability** experiment where
`alpha` becomes a function of a mixing-time parameter `t`. Small `alpha(t)`
would generate nearly pure samples near simplex corners; larger `alpha(t)`
would move samples toward increasingly mixed, interior compositions. That
extension would ask when true coordinates and compositions remain recoverable.
We expect coordinate recovery to become difficult when corner support
disappears, while composition recovery may remain useful for describing the
observed mixtures even when the true vertices are only weakly supported.

```{r helpers}
`%||%` <- function(x, y) {
    if (is.null(x)) y else x
}

sample_simplex_weights <- function(n, K, alpha = 1, corner_weights = NULL) {
    if (is.null(corner_weights)) {
        shape <- rep(alpha, K)
    } else {
        shape <- alpha * corner_weights / mean(corner_weights)
    }
    Z <- matrix(stats::rgamma(n * K, shape = rep(shape, each = n), rate = 1),
        nrow = n,
        ncol = K
    )
    Z / rowSums(Z)
}

random_archetypes <- function(K, M, seed) {
    set.seed(seed)
    A <- matrix(stats::rnorm(K * M), nrow = K, ncol = M)
    colnames(A) <- paste0("x", seq_len(ncol(A)))
    rownames(A) <- paste0("true_A", seq_len(nrow(A)))
    A
}

scenario_definitions <- function() {
    list(
        triangle_easy = list(
            name = "triangle_easy",
            title = "Easy triangle",
            K_true = 3L,
            n = 120L,
            M = 2L,
            alpha = 0.35,
            noise = 0.03,
            seed = 1001L,
            A = rbind(c(0, 0), c(1, 0), c(0.25, 0.9))
        ),
        simplex_moderate = list(
            name = "simplex_moderate",
            title = "Moderate 4-archetype simplex",
            K_true = 4L,
            n = 160L,
            M = 5L,
            alpha = 0.45,
            noise = 0.06,
            seed = 2001L
        ),
        high_dim_noisy = list(
            name = "high_dim_noisy",
            title = "Noisy high-dimensional simplex",
            K_true = 5L,
            n = 200L,
            M = 10L,
            alpha = 0.85,
            noise = 0.10,
            seed = 3001L
        ),
        imbalanced_corner = list(
            name = "imbalanced_corner",
            title = "Imbalanced sparse corner",
            K_true = 4L,
            n = 170L,
            M = 4L,
            alpha = 0.30,
            corner_weights = c(1, 1, 1, 0.18),
            noise = 0.05,
            seed = 4001L
        )
    )
}

select_scenarios <- function(budget) {
    scenarios <- scenario_definitions()
    scenarios[seq_len(min(budget$n_scenarios, length(scenarios)))]
}

make_benchmark_data <- function(scenario, data_rep, budget) {
    seed <- scenario$seed + 100L * data_rep
    set.seed(seed)
    K_true <- scenario$K_true
    M <- scenario$M
    A <- scenario$A
    if (is.null(A)) {
        A <- random_archetypes(K_true, M, seed = scenario$seed)
    } else {
        A <- as.matrix(A)
        colnames(A) <- paste0("x", seq_len(ncol(A)))
        rownames(A) <- paste0("true_A", seq_len(nrow(A)))
    }
    S <- sample_simplex_weights(
        n = scenario$n,
        K = K_true,
        alpha = scenario$alpha,
        corner_weights = scenario$corner_weights
    )
    X <- S %*% A
    X <- X + matrix(stats::rnorm(length(X), sd = scenario$noise),
        nrow = nrow(X),
        ncol = ncol(X)
    )
    colnames(X) <- colnames(A)
    rownames(X) <- paste0(scenario$name, "_", data_rep, "_", seq_len(nrow(X)))
    list(X = X, A_true = A, S_true = S, scenario = scenario, data_rep = data_rep)
}

candidate_K <- function(scenario, budget) {
    seq_len(scenario$K_true + budget$K_extra)
}

final_loss <- function(fit) {
    utils::tail(fit[["loss"]][["loss"]], 1L)
}

fit_repeated_grid <- function(benchmark_data, budget) {
    scenario <- benchmark_data$scenario
    K_values <- candidate_K(scenario, budget)
    fits <- vector("list", length(K_values))
    names(fits) <- paste0("K", K_values)

    for (i in seq_along(K_values)) {
        K <- K_values[[i]]
        fits_for_K <- vector("list", budget$n_restarts_per_K)
        for (restart in seq_len(budget$n_restarts_per_K)) {
            set.seed(scenario$seed + 10000L * benchmark_data$data_rep + 100L * K + restart)
            fits_for_K[[restart]] <- suppressWarnings(run_aa(
                benchmark_data$X,
                K = K,
                init = "random",
                max_iter = budget$max_iter,
                nrep = budget$nrep_fit
            ))
        }
        fits[[i]] <- fits_for_K
    }
    list(
        scenario = scenario,
        data_rep = benchmark_data$data_rep,
        K = K_values,
        data = benchmark_data,
        fits = fits
    )
}

best_fit_by_loss <- function(grid_result) {
    lapply(grid_result$fits, function(fits_for_K) {
        losses <- vapply(fits_for_K, final_loss, numeric(1L))
        fits_for_K[[which.min(losses)]]
    })
}

score_ic_grid <- function(grid_result) {
    best_fits <- best_fit_by_loss(grid_result)
    df_methods <- c("suleman", "covariance", "parametric", "active")
    types <- c("aic", "bic")
    rows <- vector("list", length(best_fits) * length(df_methods) * length(types))
    idx <- 0L

    for (K_name in names(best_fits)) {
        K <- as.integer(sub("^K", "", K_name))
        fit <- best_fits[[K_name]]
        for (type in types) {
            for (df_method in df_methods) {
                idx <- idx + 1L
                value <- suppressWarnings(tryCatch(
                    aa_ic(fit, type = type, df_method = df_method, n_eff = "samples"),
                    error = function(e) NA_real_
                ))
                rows[[idx]] <- data.frame(
                    scenario = grid_result$scenario$name,
                    data_rep = grid_result$data_rep,
                    K_true = grid_result$scenario$K_true,
                    K = K,
                    diagnostic = toupper(type),
                    variant = df_method,
                    value = value,
                    direction = "min",
                    stringsAsFactors = FALSE
                )
            }
        }
    }
    do.call(rbind, rows)
}

mean_pairwise_consistency <- function(fits, what, data) {
    if (length(fits) < 2L) {
        return(NA_real_)
    }
    pairs <- utils::combn(seq_along(fits), 2L)
    values <- vapply(seq_len(ncol(pairs)), function(i) {
        suppressWarnings(tryCatch(
            consistency(fits[[pairs[1L, i]]], fits[[pairs[2L, i]]], what = what, data = data),
            error = function(e) NA_real_
        ))
    }, numeric(1L))
    if (all(is.na(values))) {
        return(NA_real_)
    }
    mean(values, na.rm = TRUE)
}

score_consistency_grid <- function(grid_result) {
    what_values <- c("coordinates", "compositions")
    rows <- vector("list", length(grid_result$fits) * length(what_values))
    idx <- 0L

    for (K_name in names(grid_result$fits)) {
        K <- as.integer(sub("^K", "", K_name))
        fits_for_K <- grid_result$fits[[K_name]]
        for (what in what_values) {
            idx <- idx + 1L
            rows[[idx]] <- data.frame(
                scenario = grid_result$scenario$name,
                data_rep = grid_result$data_rep,
                K_true = grid_result$scenario$K_true,
                K = K,
                diagnostic = "consistency",
                variant = what,
                value = mean_pairwise_consistency(fits_for_K, what, grid_result$data$X),
                direction = "max",
                stringsAsFactors = FALSE
            )
        }
    }
    do.call(rbind, rows)
}

summarize_selection <- function(scores) {
    split_keys <- paste(scores$scenario, scores$data_rep, scores$diagnostic, scores$variant, sep = "\r")
    parts <- split(scores, split_keys)
    rows <- lapply(parts, function(x) {
        x <- x[is.finite(x$value), , drop = FALSE]
        if (nrow(x) == 0L) {
            return(NULL)
        }
        if (identical(x$direction[[1L]], "min")) {
            best <- x[order(x$value, x$K), ][1L, ]
        } else {
            best <- x[order(-x$value, x$K), ][1L, ]
        }
        data.frame(
            scenario = best$scenario,
            data_rep = best$data_rep,
            K_true = best$K_true,
            diagnostic = best$diagnostic,
            variant = best$variant,
            selected_K = best$K,
            recovered = best$K == best$K_true,
            stringsAsFactors = FALSE
        )
    })
    do.call(rbind, rows)
}

summarize_recovery <- function(selection) {
    split_keys <- paste(selection$scenario, selection$diagnostic, selection$variant, sep = "\r")
    parts <- split(selection, split_keys)
    rows <- lapply(parts, function(x) {
        data.frame(
            scenario = x$scenario[[1L]],
            diagnostic = x$diagnostic[[1L]],
            variant = x$variant[[1L]],
            n = nrow(x),
            K_true = x$K_true[[1L]],
            mean_selected_K = mean(x$selected_K),
            recovery_rate = mean(x$recovered),
            under_rate = mean(x$selected_K < x$K_true),
            over_rate = mean(x$selected_K > x$K_true),
            stringsAsFactors = FALSE
        )
    })
    out <- do.call(rbind, rows)
    out[order(out$scenario, out$diagnostic, out$variant), ]
}

mean_curve_scores <- function(scores) {
    split_keys <- paste(scores$scenario, scores$diagnostic, scores$variant, scores$K, sep = "\r")
    parts <- split(scores, split_keys)
    rows <- lapply(parts, function(x) {
        data.frame(
            scenario = x$scenario[[1L]],
            K_true = x$K_true[[1L]],
            K = x$K[[1L]],
            diagnostic = x$diagnostic[[1L]],
            variant = x$variant[[1L]],
            value = mean(x$value, na.rm = TRUE),
            stringsAsFactors = FALSE
        )
    })
    do.call(rbind, rows)
}

plot_selection_curves <- function(scores, scenario_name, diagnostic, main = NULL) {
    cur <- scores[scores$scenario == scenario_name & scores$diagnostic == diagnostic, , drop = FALSE]
    cur <- mean_curve_scores(cur)
    cur <- cur[is.finite(cur$value), , drop = FALSE]
    if (nrow(cur) == 0L) {
        plot.new()
        title(main = main %||% diagnostic)
        text(0.5, 0.5, "No finite values")
        return(invisible(cur))
    }

    variants <- unique(cur$variant)
    colors <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7")
    colors <- colors[seq_along(variants)]
    ylim <- range(cur$value, finite = TRUE)
    plot(
        NA,
        xlim = range(cur$K),
        ylim = ylim,
        xlab = "Number of archetypes (K)",
        ylab = diagnostic,
        main = main %||% diagnostic,
        frame.plot = FALSE
    )
    abline(v = cur$K_true[[1L]], lty = 2, col = "gray45")
    for (i in seq_along(variants)) {
        one <- cur[cur$variant == variants[[i]], , drop = FALSE]
        one <- one[order(one$K), , drop = FALSE]
        lines(one$K, one$value, type = "b", pch = 19, col = colors[[i]])
    }
    legend("topright", legend = variants, col = colors, pch = 19, lty = 1, bty = "n")
    invisible(cur)
}
```

The scenarios selected by the active budget are:

```{r scenarios}
scenarios <- select_scenarios(benchmark_budget)
scenario_table <- data.frame(
    scenario = vapply(scenarios, `[[`, character(1L), "name"),
    K_true = vapply(scenarios, `[[`, integer(1L), "K_true"),
    n = vapply(scenarios, `[[`, integer(1L), "n"),
    M = vapply(scenarios, `[[`, integer(1L), "M"),
    noise = vapply(scenarios, `[[`, numeric(1L), "noise"),
    stringsAsFactors = FALSE
)
scenario_table
```

## Run the Benchmark

The next chunk is the expensive part. In `development` mode it runs only a
small subset. In `article` mode it runs all scenarios, multiple datasets per
scenario, and multiple repeated fits per candidate `K`.

```{r run-benchmark}
grid_results <- list()
idx <- 0L
for (scenario in scenarios) {
    for (data_rep in seq_len(benchmark_budget$n_reps_data)) {
        idx <- idx + 1L
        benchmark_data <- make_benchmark_data(scenario, data_rep, benchmark_budget)
        grid_results[[idx]] <- fit_repeated_grid(benchmark_data, benchmark_budget)
    }
}
```

```{r score-benchmark}
#| column: page


ic_scores <- do.call(rbind, lapply(grid_results, score_ic_grid))
consistency_scores <- do.call(rbind, lapply(grid_results, score_consistency_grid))
all_scores <- rbind(ic_scores, consistency_scores)

selection <- summarize_selection(all_scores)
recovery_summary <- summarize_recovery(selection)
recovery_summary
```

## Scenario Curves

Each panel marks the true `K` with a vertical reference line. IC curves should
be read by their minimum. Consistency curves should be read by their maximum.

```{r scenario-curves, fig.width = 10, fig.height = 4.5, results = "hide"}
#| column: page


for (scenario in scenarios) {
    op <- par(mfrow = c(1, 3), mar = c(4, 4, 3, 1), bty = "n")
    plot_selection_curves(all_scores, scenario$name, "AIC",
        main = paste(scenario$title, "AIC")
    )
    plot_selection_curves(all_scores, scenario$name, "BIC",
        main = paste(scenario$title, "BIC")
    )
    plot_selection_curves(all_scores, scenario$name, "consistency",
        main = paste(scenario$title, "consistency")
    )
    par(op)
}
```

## Interpretation

The benchmark separates three related but distinct questions:

- Does a criterion select the known generating `K_true`?
- Does it under-select stable but too-simple models?
- Does it over-select because extra archetypes improve the empirical hull or
  fit noise?

Information criteria and consistency can disagree because they measure
different properties. AIC and BIC penalize model size through degrees of freedom.
Coordinate consistency checks whether repeated fits place archetypes in similar
regions of input space. Composition consistency checks whether samples receive
similar archetypal mixtures across repeated fits.

High consistency is not automatically evidence that `K` is correct. A
one-archetype model can be very stable because there are few choices to make.
Conversely, a correct but hard geometry may have lower consistency if repeated
fits find different local optima. In practice, consistency is best interpreted
alongside IC curves, reconstruction loss, restarts, and domain knowledge.

## Extending the Benchmark

The benchmark scaffold is intentionally modular. Future comparisons can add
criteria by extending the scoring step rather than changing the data generator.
For example, a group-lasso selector could be added as another `diagnostic` with
its own `variant` values and a declared selection direction. The existing
scenario definitions, candidate `K` grids, and recovery summaries would still
apply.

The most useful additions would be:

- repeated simulations with larger `n_reps_data`;
- stronger noise and missing-data scenarios;
- sparse or regularized fitters whose active sets make `df_method = "active"`
  especially relevant;
- runtime summaries, because consistency-based selection is substantially more
  expensive than scoring one best fit per `K`.
